\documentclass[11pt, reqno]{amsart}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}

\usepackage{showlabels}

\usepackage{tikz}
\usetikzlibrary{arrows}

\usepackage{lipsum}

\usepackage{listings}

\usepackage{enumerate}
\usepackage{enumitem}

\usepackage{tabu}

\lstset{basicstyle=\ttfamily, mathescape=true}

\newcommand{\R}{\mathbb{R}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\HH}{\mathcal{H}}
\newcommand{\BB}{\mathcal{B}}

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{prop}{Prop}

\newtheorem{definition}{Definition}

\newtheorem{conjecture}{Conjecture}

\newtheorem{remark}{Remark}

\DeclareMathOperator{\dig}{dig}
\DeclareMathOperator{\intr}{int}


\DeclareMathOperator{\len}{le}
\newcommand{\LE}{\mathbf{LE}}

\title[\textbf{Optimal bounds for $b$-ary measures}]{\textbf{On optimal bounds for comparing Binary (and $b$-ary) Nest Measures with the Hausdorff Measure on $\R$}}


\author{Duarte Maia}
\address{Department of Mathematics, Instituto Superior Técnico}
\email{duarte.nascimento@tecnico.ulisboa.pt}

\iffalse

\author{Jorge Drumond Silva}
\address{Department of Mathematics, Instituto Superior Técnico}
\email{jsilva@math.ist.utl.pt}

\fi

\date{}


\begin{document}


\begin{abstract}
This is a test.
\end{abstract}

\maketitle


\section{Introduction}

Hausdorff measure is a central topic in geometric measure theory. It is a generalization of the concept of length, area and volume for arbitrary Metric spaces and possibly nonintegral dimensions. It was introduced by Felix Hausdorff (1919) as a generalization of a construction by Carathéodory (1914), and important applications have since been found in diverse areas, such as fractal geometry, Brownian motion \cite{brownian} and harmonic analysis \cite{fefferman}.

The Hausdorff measure is actually a family of outer measures, denoted by $\HH_s$ where $s$ is a nonnegative real parameter. For integral $s$, these coincide (modulo a constant factor) with the usual Lebesgue measure on $\R^n$. We refer to $\HH_s$ as \emph{the $s$-dimensional Hausdorff measure}.

If one were to fix a set $E$ and plot the graph of $\HH_s(E)$ as a function of $s$, one would find that this graph has a very simple appearance. Indeed, $\HH_s(E)$ would, for small enough $s$, be equal to $\infty$, followed by a sudden and sharp drop to 0. The value of $s$ where this drop occurs called the \emph{Hausdorff dimension of $E$}. It measures roughly how `space-filling' $E$ is. So, for example, a square occupies 2-dimensional space, a sphere occupies 3-dimensional space, and, for a slightly more exotic example, the standard Cantor `middle thirds' set `occupies $\log_3 2$-dimensional space'.

To aid in the study of Hausdorff dimension, Besicovitch (1952) introduced the notion of \emph{binary measure}, which is a modification of Hausdorff measure that, while not exactly equivalent, nevertheless preserves dimension. That is, the dimension of a set from the perspective of the Hausdorff measure and the binary measure is the same. The reason for this is an inequality of the form
\begin{equation}\label{hbineq1}
\text{Hausdorff measure} \leq \text{Binary measure} \leq C \times \text{Hausdorff measure},
\end{equation}
which ensures that both agree where they are either zero or infinity.

In this article, we intend to investigate inequality \eqref{hbineq1}.

We begin in section \ref{sectgenbmeasures} by generalizing the notion of binary measures to the so-called $b$-ary measures, where $b$ represents a natural number greater than one. We denote the $s$-dimensional $b$-ary measure by $\BB_{bs}$, and show an inequality similar to \eqref{hbineq1}. In particular, we show that, for any set $E$,
\begin{equation}\label{hbineq2}
\HH_s(E) \leq \BB_{bs}(E) \leq (b+1) \HH_s(E).
\end{equation}

In section \ref{sec5} we investigate tools to sharpen this bound, and in sections \ref{sec6} and \ref{sec7} use these tools to draw a connection between the $b$-ary measures and the theory of non-integral digit expansions. In particular, we show that we can replace the $b+1$ in inequality \eqref{hbineq2} by a constant $K$, defined as the smallest constant such that a certain function $g(t)$ can be bounded from above by $K t^s$.

In sections \ref{calck} and \ref{optimization} we introduce a method to estimate the numerical value of this $K$ to arbitrary precision.

Turning to lower bounds, in section \ref{seclowerbound} we find a lower bound to the value by which $b+1$ may be replaced in equation \eqref{hbineq2}. In other words, we construct an example of a set where the binary and the Hausdorff measure differ significatively, again using the tools of non-integral digit expansions.

Finally, in section \ref{secfinal} we use these upper and lower bounds to construct a set $E$ satisfying, for $s = 1/2$, $\BB_{5s}(E) < \BB_{4s}(E)$. Stronger still: if we let $G = \frac{\BB_{4s}(E)}{\HH_s(E)}$, there exsts $k < G$ such that for all $F$ we have $\BB_{5s}(F) \leq k \times \HH_s(F)$. This shows that there exist dimensions $s$ such that $\BB_{5s}(F)$ is a better estimate for the Hausdorff measure of $F$ than $\BB_{4s}(F)$, which, in light of inequality \eqref{hbineq2}, is somewhat unintuitive, as one would expect the inequality to get less and less sharp as $b$ increases.


\section{Hausdorff and Binary measures: a review}

\subsection{Hausdorff}

Let $X$ be a metric space, with distance $d$. We define the \emph{$s$-dimensional Hausdorff measure on $X$} as follows:

First, for $E \subseteq X$, $s \in \R^+_0$ and $\delta > 0$, define $\HH_s^\delta(E) := \inf \sum_i \lvert U_i \rvert^s$, where:

\begin{itemize}

\item $\lvert E \rvert$ denotes the \emph{diameter of $E$}, defined by the $\sup$ of distances between elements of $E$;\footnote{We follow the convention that the diameter of the empty set is 0.}

\item The infimum is taken over countable coverings $\{U_i\}_{i \in \N}$ of $E$ such that the diameter of all $U_i$ is less than $\delta$. This is usually abbreviated to `$U$ is a \emph{$\delta$-covering} of $E$'.

\end{itemize}

As $\delta$ decreases this quantity increases. Hence, the limit $\lim_{\delta \to 0} \HH_s^\delta(E)$ exists in $\left[ 0, \infty \right]$. It is this quantity that is called the \emph{$s$-dimensional Hausdorff measure of $E$}:

\begin{definition}(Hausdorff Measure) Let $E \subseteq X$ and $s \in \R^+_0$. Define the $s$-dimensional Hausdorff measure of $E$ as:

 \[\HH_s(E) := \lim_{\delta \to 0} \HH_s^\delta(E).\]
\end{definition}

\subsection{Dimension}

Given a subset $E$ of a metric space $X$, there exists $\sigma \in \left[0, \infty\right]$ such that:

\begin{itemize}
\item For $s < \sigma$, $\HH_s(E) = \infty$;

\item For $s > \sigma$, $\HH_s(E) = 0$.
\end{itemize}

To prove this, we first show the following lemma:

\begin{lemma}\label{helperdimension}
Let $t < s$ be real nonnegative numbers. Then, if $\HH_t(E) < \infty$ we have $\HH_s(E) = 0$.
\end{lemma}

\begin{proof}
Suppose $t < s$ and $\HH_t(E) < \infty$. Put $M = \HH_t(E) + 1$. Then, for any $\delta > 0$ we may find a $\delta$-covering $\{U_i\}$ of $E$ such that $\sum \lvert U_i \rvert^t < M$. Hence, 

\begin{align*}
\sum \lvert U_i \rvert^s &= \sum \lvert U_i \rvert^t \lvert U_i \rvert^{s - t}\\
&\leq \sum \lvert U_i \rvert^t \delta^{s-t}\\
&\leq \delta^{s-t} M.
\end{align*}

Therefore, $\HH_s^\delta(E) \leq \delta^{s-t} M$, and so making $\delta \to 0$ we get $\HH_s(E) = 0$, as desired.
\end{proof}

We can now define the dimension of a set $E$.

\begin{definition} (Hausdorff dimension of a set) Let $E \subseteq X$. We define the dimension of $E$ as:

\[\sigma_E := \inf \{\, s \mid \HH_s(E) < \infty \,\}.\]
\end{definition}

\begin{prop}
Let $E \subseteq X$, where $X$ is a metric space. Then, the dimension of $E$ is such that the Hausdorff measure of $E$ is null for dimensions higher than it, and infinite for dimensions lower than it. Symbolically:

\begin{itemize}
\item For $s < \sigma_E$, $\HH_s(E) = \infty$;

\item For $s > \sigma_E$, $\HH_s(E) = 0$.
\end{itemize}

\end{prop}

\begin{proof}
The first inequality is trivial by definition of $\sigma_E$.

To show the second inequality, notice that if $s > \sigma_E$ then there exists $t < s$ such that $\HH_t(E) < \infty$. By lemma \ref{helperdimension}, we have $\HH_s(E) = 0$, as desired.
\end{proof}

This shows that the Hausdorff dimension of a set is the only dimension for which its corresponding Hausdorff measure may be nontrivial, that is, different from $0$ or $\infty$.

\subsection{Net measures}

Hausdorff measure, and by extension Hausdorff dimension, is often very difficult to compute precisely as its definition requires one to look at all possible $\delta$-coverings. As such, there is interest in finding definitions that would allow simpler methods of computing, or at least estimating, these quantities.   

We will restrict our attention to $\R^n$, and later on to $\R$, as a lot of what follows is not immediately generalizable to other metric spaces.

First, notice that it is not uncommon to restrict one's attention to specific coverings. Indeed, if in the definition of $\HH_s^\delta$ one takes the infimum over $\delta$-coverings \emph{by open sets} instead of arbitrary ones, the value of $\HH_s^\delta$ may change, but the limit as $\delta \to 0$ remains the same. Hence, one could consider the Hausdorff measure defined by using only open $\delta$-coverings with no loss of generality.

The notion of net measure arises from the idea of redefining the Hausdorff measure with a more manageable collection of sets to use for coverings. In the context of $\R^n$, Besicovitch (1952) used the so-called binary intervals, which is a construction we will make solely for the case of $n = 1$, but more exhaustive treatments can be found in \cite{falconer, rogers}.

\begin{definition}
Define the so-called \emph{binary intervals} in $\R$ as the set of intervals of the form $\left[ k 2^j, (k+1) 2^j \right[$, for $k, j \in \Z$.

The binary measure is defined analogously to the Hausdorff measure, except instead of arbitrary coverings we consider only $\delta$-coverings by binary intervals. It is denoted by $\BB_s(E) = \lim_{\delta \to 0} \BB_s^\delta(E)$.
\end{definition}

The usefulness of this definition lies in that, even though these measures do \emph{not} coincide in general with the Hausdorff measure, one can find bounds between them.

The following theorem is found in \cite{falconer, rogers} and is stated here without proof, though in the sequence of this document we will prove stronger results for the particular case of $n = 1$.

\begin{theorem}\label{badbound}
Consider the Hausdorff and Binary measures defined on $\R^n$. For any set $E$, we have

\begin{equation}\label{badboundeq}
\HH_s(E) \leq \BB_s(E) \leq 3^n 2^{n^2} \HH_s(E).
\end{equation}
\end{theorem}

\begin{remark}
Even though the theorem \ref{badbound} will not be proved here, it should be noticed that one of the inequalities is trivial.

Indeed, for any $\delta$ we have $\HH_s^\delta(E) \leq \BB_s^\delta(E)$, since both are infima taken over coverings of $E$, but the set of coverings considered in the definition of the Hausdorff measure is a contains all of those considered in the binary measure. Making $\delta \to 0$, we have inequality $\HH_s(E) \leq \BB_s(E)$.
\end{remark}

Theorem \ref{badbound} has the following consequence: defining the ``binary dimension'' of a set analogously to how we defined the Hausdorff dimension, these two concepts would coincide. Indeed, the Hausdorff measure of a set is zero iff the binary measure is zero, and the same holds for when this measure is infinite.

Hence, if we are interested in the Hausdorff dimension of a set, we may begin by investigating its binary measure.

Notice, however, that the provided bound is not very sharp. Indeed, for $n = 1$ the constant given in inequality \eqref{badboundeq} amounts to 6, but we will see below that it can be significantly lowered without much effort. (And then lowered even more, with significantly more effort.)

Of course, this is of no consequence for dimensional considerations. However, the results that follow allow for much sharper bounds, and hopefully better understanding of how these two relate.

\subsection{Hausdorff Coverings}

As mentioned above, it is sometimes convenient for one to only consider open coverings of sets instead of allowing for the full generality of any possible covering. This is not the only simplification that will be useful for us to do.

Notice that any monotone operation (that is, $f(E) \supseteq E$) that can be done on a set without increasing its diameter can be assumed without loss of generality.

To use an example: taking the closure of a set $U$ does not increase its diameter. As such, given any $\delta$-covering $\{U_i\}$ of a set $E$, the collection $\{ \overline{U_i} \}$ is also a $\delta$-covering of $E$, with all sets closed, and with all diameters the same, whence the Hausdorff sum is also the same. This shows, then, that we can assume without loss of generality that the infimum in the definition of $\HH_s^\delta$ can be taken over coverings by closed sets.

A more specific example, useful for our purposes, is (in $\R^n$) the operation of taking the convex hull, which satisfies the condition of monotony without increasing diameters. This shows, then, that we may take infima over convex sets with no loss of generality.

In the particular case of $\R$, what this shows is that we can assume that the sets we are using for coverings are intervals. Sometimes we will additionally assume that these intervals are closed, and sometimes we will instead assume them to be open. In any case, these hypotheses harm in no way the generality of the results obtained.

As a final remark, it is often useful to exclude, in our coverings, cases where the sets have diameter 0. Indeed, many authors explicitly exclude the case of 0-diameter sets in the definition of $\delta$-coverings.

\section{Better elementary bounds}

We have already shown that, in $\R$, $\HH_s \leq \BB_s$. We will now proceed to establish a bound in the opposite direction, beginning our journey to find optimal bounds.

Everything that follows will be done in the context of $\R$.

\begin{theorem} For all $s \in \R^+_0$ the binary measure and Hausdorff measure obey the inequality

\begin{equation}\label{easybound}
\BB_s \leq 3 \HH_s
\end{equation}
\end{theorem}

\begin{proof}
To prove this fact, we will show that for any $E$ and $\delta$ we have the inequality $\BB_s^\delta(E) \leq 3 \HH_s^\delta(E)$. To this effect, we will show that given any $\delta$-covering of $E$ we can construct a binary $\delta$-covering of $E$ such that its `Hausdorff sum' is at most three times bigger than the original's.

Fix, then, a $\delta$-covering $\{U_i\}$ of $E$, which we suppose without loss of generality to be composed of intervals with diameter strictly greater than zero. Define, for each $i$, $j_i$ to be the greatest integer such that $2^{j_i} < \lvert U_i \rvert$, and hence $2^{j_i} < \delta$ for all $i$.

Consider how many intervals of the form $\left[ k 2^{j_i}, (k + 1) 2^{j_i} \right[$ intersect $U_i$. There are at most three, and their union definitely covers $U_i$.

Now, consider the covering $\{I_i\}$ obtained from $\{U_i\}$ by replacing each $U_i$ by the intervals mentioned above. We now have a binary $\delta$-covering of $E$, with Hausdorff sum equal to
\[\sum \lvert I_i \rvert^s \leq 3 \sum \lvert U_i \rvert^s.\]

With this inequality holding because for every $U_i$ we assign its three corresponding intervals, each with diameter less than $\lvert U_i \rvert$. This completes our proof, since
\begin{align*}
3 \HH_s^\delta &= 3 \times \inf \text{over $U_i$}\\
&\geq \inf \text{over the corresponding $I_i$}\\
&\geq \inf \text{over all binary $\delta$-coverings}\\
&= \BB_s^\delta
\end{align*}

\end{proof}

This bound, while twice as sharp as the one mentioned in theorem \ref{badbound}, can still be significantly improved. The proof, however, touches on what will be our main tool to try to control the bound. The idea is to find a way to replace any $\delta$-covering with a binary $\delta$-covering such that, if the first covering has Hausdorff sum $A$, the latter has Hausdorff sum at most $CA$, where $C$ is the bound on $\frac \BB \HH$ we desire to prove. Taking this a step further, we can focus merely on finding a way to, given an interval of length $\ell$, covering it by binary intervals of lengths $m_i$ such that $\sum m_i^s \leq C \ell^s$. One might have concerns about making sure that these binary intervals still form a $\delta$-covering, but that is not an issue, as we will see in the following lemma:

\begin{lemma}
Fix a dimension $s \in \R^+$ and suppose $C \in \R^+$ is such that, for any interval $I$ of length $\ell$, we can find a countable binary covering of $I$ by intervals $\{J_i\}$ of lengths $m_i$ such that
\[\sum m_i^s \leq C \ell^s.\]

Then $\BB_s \leq C \HH_s$.
\end{lemma}

\begin{proof}
This proof proceeds much like the previous one, with one caveat.

Given some $E$, let $I_i$ be any $\delta$-covering of $E$ by intervals. For each $I_i$, let $\{J_{ij}\}_{j \in \N}$ be a collection of binary intervals such that  $\sum_j \lvert J_{ij} \rvert^s \leq C \lvert I_i \rvert^s$.

First, notice that, while it is \emph{not} necessarily true that the collection $\{J_{ij}\}_{i,j \in \N}$ forms a $\delta$-covering, we can be assured that it forms at most a $C^{1/s}\delta$-covering, as, for any $i, j$ we have
\[ \lvert J_{ij} \rvert^s \leq \sum_j \lvert J_{ij} \rvert^s \leq C \lvert I_i \rvert^s \leq C \delta^s,\]
and so $\lvert J_{ij} \rvert \leq C^{1/s} \delta$, as desired.

The argument now follows much like the previous one, except we can only show that $\BB_s^{C^{1/s} \delta} \leq C \HH_s^\delta$. This turns out to be enough, however, as taking $\delta \to 0$ also yields $C^{1/s} \delta \to 0$, and so both sides of this inequality converge to $\BB_s$ and $\HH_s$, respectively.
\end{proof}

\begin{remark} \label{dimnotzero}
We had to exclude from consideration the case $s = 0$ in this last proposition, as we will have to in a lot of what follows. This is a nonissue, however, as this particular case is extremely easy to examine. Indeed, for this dimension both the Hausdorff and the binary measure turn out to be simply the counting measure on $\R$.

As such, in all that follows, unless otherwise specified, \textbf{it should be assumed that $s \neq 0$}.
\end{remark}

This previous lemma will be central in what follows. Indeed, our strategy to find a bound of the form $\BB_s \leq C \HH_s$ will be to, given any interval $I$, find a binary covering $\{J_i\}$ of $I$ such that $\sum \lvert J_i \rvert^s \leq C \lvert I \rvert^s$. But before further pursuing this idea, we introduce a generalization of the binary measure.

\section{$b$-ary measures}\label{sectgenbmeasures}

As mentioned before, for the purposes of dimensional analysis, the binary measures are enough. However, for the purposes of estimating the Hausdorff measure, it might be worth looking into something slightly more general than binary intervals.

\begin{definition}
Define the $b$-ary intervals, where $b \in \N_{\geq 2}$, as those of the form
\[\left[k b^j, (k+1) b^j \right[, \;k, j \in \Z.\]

The definition for the $s$-dimensional $b$-ary measure, denoted $\BB_{bs}$, and the corresponding $\delta$-measures $\BB_{bs}^\delta$, should be evident by comparison to the definition of the binary one, using the $b$-ary intervals instead of the binary ones.
\end{definition}

The results of the last section can be translated into this slightly more general context without much added effort. We state the results here without proof.

\begin{lemma}
Fix a dimension $s \in \R^+$, a base $b \in \N_{\geq 2}$, and suppose $C \in \R^+$ is such that, for any interval $I$ of length $\ell$, we can find a countable $b$-ary covering of $I$ by intervals $\{J_i\}$ of lengths $m_i$ such that
\[\sum m_i^s \leq C \ell^s.\]

Then $\BB_{bs} \leq C \HH_s$.

In particular, one can easily show the premise is true for $C = b+1$, whence one gets the bound
\[\HH_s \leq \BB_{bs} \leq (b+1) \HH_s.\]
\end{lemma}

This last bound hints at something we will study: it appears that, the bigger $b$ is (that is, the more pieces we divide our intervals in as $j$ decreases) the worse the estimate. This makes sense, since as $b$ increases there is less we can discern because since there is a larger discrepancy between the size of one layer and the next we lose the ability to gauge more precisely the intermediate sizes.

This intuition proves to be partially correct, as the following proposition shows:

\begin{prop}
Let $b, c \in \N_{\geq 2}$ and suppose $c$ is a power of $b$. Then, $\BB_{bs} \leq \BB_{cs}$.
\end{prop}

\begin{proof}
This proof is trivial. Simply notice that, if $c$ is a power of $b$, any $c$-ary interval is also a $b$-ary one. As such, when taking the infima for $\BB_{bs}^\delta$ one has more sets than in $\BB_{cs}^\delta$, whence $\BB_{bs}^\delta \leq \BB_{cs}^\delta$ and the result follows.
\end{proof}

We would like to find an criterion to, given two numbers $c, b \in \N_{\geq 2}$, tell if it is the case that $\BB_{bs} \leq \BB_{cs}$. Unfortunately, we have, as of yet, been unable to find one. We do, however, have some evidence that shows that the following proposition is \textbf{false}: 

\begin{conjecture}
If $c \geq b$ then $\BB_{bs} \leq \BB_{cs}$.
\end{conjecture}

The counterexample to this possible conjecture is found closer to the end of this article.

\section{Minimal interval coverings}\label{sec5}

We focus now on the following problem: given a dimension $s$ and an interval $I$, find a $b$-ary covering of $I$ with as small a Hausdorff sum as possible. We assume $I$ to be of the form $\left[ a, b \right[$ with $a < b$, as we can always assume our Hausdorff coverings are composed of intervals of this kind. The reason we are assuming this is so that this plays well with our $b$-ary intervals, which are also half-open.

In truth, what we are actually concerned about is finding as small a $b$-ary covering of an interval $I$ as possible. In particular, we want to determine the smallest $C$ such that for any interval of length $\ell$ we can find a binary covering with Hausdorff sum less than or equal to $C \ell^s$. This can be written as
\begin{equation}\label{defc}
C = \sup \frac{\BB_{bs}^\infty(I)}{\lvert I \rvert^s}
\end{equation}
where the supremum is taken over the half-open nonempty intervals $I$ and the usage of $\delta = \infty$ is intended to imply that any $b$-ary covering is admissible.

Notice, now, that an interval is uniquely determined by its length and its position. We can, then, write \eqref{defc} as
\[\sup_\ell \sup_\text{$I$ of length $\ell$} \frac{\BB_{bs}^\infty(I)}{\ell^s},\]
which is the same as
\[\sup_\ell \frac{\sup_I \BB_{bs}^\infty(I)}{\ell^s}.\]

We investigate, then, $\sup \BB_{bs}^\infty(I)$, where the supremum is taken over intervals of a fixed length $\ell$.

We will now show that we can assume, in addition, that our interval contains the origin. To this effect, given any interval $I$ we will construct an interval $I'$ of same length containing the origin, such that $\BB_{bs}^\infty(I') \geq \BB_{bs}^\infty(I)$.

This construction requires splitting into cases.

Let $\ell = \lvert I \rvert$, and $j$ such that $b^j < \ell \leq b^{j+1}$.

\begin{enumerate}[label=\textbf{Case \arabic*:}]

\item \textbf{There exists $k$ such that $k b^{j+1} \in I$.}

In this event, let $I'$ be the interval given by translating $I$ by $-k b^{j+1}$. Two easy observations are in order.

First, notice that
\[\BB_{bs}^\infty(I') = \BB_{bs}^{b^{j+1}}(I'),\]
because any covering of $I'$ using an interval of size greater than $b^{j+1}$ can be `improved' by replacing it with a smaller interval. Furthermore, we have
\begin{equation}\label{transinvariance}
\BB_{bs}^{b^{j+1}}(I') = \BB_{bs}^{b^{j+1}}(I),
\end{equation}
because the translation done to turn $I$ into $I'$ preserves binary intervals of size $b^{j+1}$ or less.

These two, together with the fact that $\BB_{bs}^{b^{j+1}}(I) \geq \BB_{bs}^\infty(I)$, give us the desired inequality
\[\BB_{bs}^\infty(I') \geq \BB_{bs}^\infty(I).\]

\item \textbf{There exists an interval of length $b^{j+1}$ within which $I$ is contained.}

In this event, 
\[\BB_{bs}^\infty(I) = \min(\BB_{bs}^{b^j}(I), b^{(j+1)s}).\]

We know for sure there exists at least one point of the form $k b^j$ in $I$. Let $I'$ be given by translating $I$ by $-k b^j$. We proceed to compare $\BB_{bs}^\infty(I')$ and $\BB_{bs}^\infty(I)$.

\begin{enumerate}[label=\textbf{Subcase \alph*:}]

\item \textbf{If} $\BB_{bs}^{b^j}(I) \leq b^{(j+1)s}$, by the same argument used to deduce equation \eqref{transinvariance}, we have
\[\BB_{bs}^{b^j}(I') = \BB_{bs}^{b^j}(I) \leq b^{(j+1)s},\]
and thus $\BB_{bs}^{b^j}(I') = \BB_{bs}^\infty(I')$, as any coverings containing intervals bigger then $b^j$ can be safely ignored. Finalizing, we get
\[\BB_{bs}^\infty(I') = \BB_{bs}^{b^j}(I') = \BB_{bs}^{b^j}(I) = \BB_{bs}^\infty(I).\]

\item \textbf{If} $\BB_{bs}^{b^j}(I) \geq b^{(j+1)s}$, simply observe that either
\[\BB_{bs}^\infty(I') = \BB_{bs}^{b^j}(I'),\]
in which case $\BB_{bs}^\infty(I') = \BB_{bs}^{b^j}(I) \geq \BB_{bs}^\infty(I)$, as desired.

Or it could happen that
\[\BB_{bs}^\infty(I') < \BB_{bs}^{b^j}(I'),\]
in which case we can safely assert $\BB_{bs}^\infty(I') \geq b^{(j+1)s} = \BB_{bs}^\infty(I)$ since we need only consider coverings that contain at least one interval of length $b^{j+1}$ and so have Hausdorff sum of at least $b^{(j+1)s}$.
\end{enumerate}
\end{enumerate}

\qed

With this construction done, we have now shown that our problem can be reduced to determining
\[\sup_{\substack{\text{$I$ half-open interval}\\ 0 \in I}} \frac{\BB_{bs}^\infty(I)}{\lvert I \rvert^s}.\]

A simple observation about the origin is that any $b$-ary interval is either to the right or to the left of the origin, but no interval touches both sides. Therefore, we can consider our covering of the left side to be independent of our covering of the right side, leading quickly to the conclusion that, if $I = \left[ a, b \right[$ with $a \leq 0 \leq b$, we have
\[\BB_{bs}^\delta(\left[a, b\right[) = \BB_{bs}^\delta(\left[a, 0\right[) + \BB_{bs}^\delta(\left[0, b\right[)\]
for any $\delta \in \left]0, \infty \right]$.

Another observation, which we will soon justify rigorously (see lemma \ref{sidedoesntmatter}), is that $\BB_{bs}^\delta(\left[a, 0\right[) = \BB_{bs}^\delta(\left[0, a\right[)$. This has the following consequence: if we define for $t \geq 0$ the function $g_{bs}(t) := \BB_{bs}^\infty(\left[0, t \right[)$, our problem becomes one of optimization:

\[ \sup_{\substack{\ell,r \in \R^+_0\\\ell+r \neq 0}} \frac{g_{bs}(\ell) + g_{bs}(r)}{(\ell + r)^s} \]

And so, we proceed to study the function $g_{bs}$, (we will often suppress the subscript when it is obvious from context and not particularly relevant) which has a connection to the theory of $b$-ary expansions of real numbers. Before getting into details, however, we prove the property mentioned above:

\begin{lemma} \label{sidedoesntmatter}
\[\BB_{bs}^\delta(\left[a, 0\right[) = \BB_{bs}^\delta(\left[0, a\right[)\]
\end{lemma}

\begin{proof}
Fix any $b$-ary $\delta$-covering of $\left[a, 0\right[$. Denote it by $\{\left[x_i, y_i\right[\}$. Consider now the collection $\{\left[-y_i, -x_i\right[\}$. 

This collection almost forms a $b$-ary $\delta$-covering of $\left[0, a\right[$. Indeed, the only points that might not be covered are those of the form $-x_i$. This can, however, be remedied by augmenting our collection with intervals of the form $\left[ -x_i, -x_i + \varepsilon_i \right[$ with arbitrarily small $\varepsilon_i$, and so reach the conclusion that there are coverings of $\left[0, a\right[$ with Hausdorff sum arbitrarily close to our original covering's, concluding that $\BB_{bs}^\delta(\left[a, 0\right[) \geq \BB_{bs}^\delta(\left[0, a\right[)$. A similar argument will show the opposite inequality.
\end{proof}

\section{$b$-ary digit expansions}\label{sec6}

In the previous section we showed that a possible approach to our problem goes through investigating the optimization problem:

\[ \sup_{\substack{\ell,r \in \R^+_0\\\ell+r \neq 0}} \frac{g_{bs}(\ell) + g_{bs}(r)}{(\ell + r)^s} \]

By itself, this reduction would not be much help. As it turns out, however, one can draw a connection between this function $g$ and the theory of noninteger base expansions.

\begin{remark}
In remark \ref{dimnotzero} we mentioned that we will be tacitly assuming $s \neq 0$. We will outline here another such assumption.

As $\R$ has Hausdorff dimension 1, for $s > 1$, both the binary and the Hausdorff measre of \emph{any} set in $\R$ will be zero. Then, the question of the relationship between these two is trivial and uninteresting. Therefore,\textbf{from now on, we always assume $s \leq 1$.}
\end{remark}

Let us now fix some $t > 0$, and investigate how one might go about calculating $g(t)$ given a dimension $s$ and a base $b$. (The case $t = 0$ is trivial, and thus not considered.) We remind the reader of the definition of $g$:

\[g_{bs}(t) := \BB_{bs}^\infty(\left[0, t \right[)\]

First, fix $b^k$ as the least integer power of $b$ that is $\geq t$. Then, a possible covering of $\left[0, t \right[$ would be the interval $\left[0, b^k\right[$. Hence, $g(t) \leq b^{ks}$.

Now, observe that $b^{k-1} < t$, and so there is a maximal integer, which we will call $d_1$, such that $d_1 b^{k-1} \leq t$. This gives us another possible covering of our interval, given by
\[\left[0, b^{k-1} \right[, \left[b^{k-1}, 2 b^{k-1} \right[, \cdots, \left[d_1 b^{k-1}, (d_1 + 1) b^{k-1} \right[,\]
which yields another upper bound, $g(t) \leq (d_1 + 1) b^{k-1}$.

We can continue this process. If we let $d_2$ be maximal such that $d_1 b^{k-1} + d_2 b^{k-2} \leq t$. The following sequence of intervals will cover $\left[0, t \right[$:
\begin{multline*}
\left[0, b^{k-1} \right[, \left[b^{k-1}, 2 b^{k-1} \right[, \cdots, \left[(d_1 - 1) b^{k-1}, d_1 b^{k-1} \right[ \\
\left[d_1 b^{k-1}, d_1 b^{k-1} + b^{k-1} \right[, , \cdots, \left[d_1 b^{k-1} + d_2 b^{k-2}, d_1 b^{k-1} + (d_2 + 1) b^{k-2} \right[,
\end{multline*}
giving us a new upper bound $d_1 b^{k-1} + (d_2 + 1) b^{k-2}$.

We will make this process more rigorous shortly. Below is a diagram exemplifying it for $b = 3$ and $t = 0.86$.


\begin{tikzpicture}
\draw[thick, {|-|}] (0,0) node[above]{0} -- (10,0) node[above]{1};
\draw[thick, {->}] (10,0) -- (12, 0) node[above]{$x$};

%The following was automatically generated with the help of the code at binary.py
\draw[thick, gray, {[-)}] (0, -1.0) -- (10, -1.0);
\draw (10, -1) node[right]{$3^0$};

\draw[thick, red, {[-)}] (0, -1.5) -- (3.333333333333333, -1.5);
\draw[thick, red, {[-)}] (3.333333333333333, -1.5) -- (6.666666666666666, -1.5);
\draw[thick, red, {[-)}] (6.666666666666666, -1.5) -- (10.0, -1.5);
\draw (10, -1.5) node[right]{$3 \times 3^{-1}$};

\draw[thick, red, {[-)}] (0, -2.0) -- (3.333333333333333, -2.0);
\draw[thick, red, {[-)}] (3.333333333333333, -2.0) -- (6.666666666666666, -2.0);
\draw[thick, orange, {[-)}] (6.666666666666666, -2.0) -- (7.777777777777777, -2.0);
\draw[thick, orange, {[-)}] (7.777777777777777, -2.0) -- (8.88888888888889, -2.0);
\draw (8.89, -2) node[right]{$2 \times 3^{-1} + 2 \times 3^{-2}$};

\draw[thick, red, {[-)}] (0, -2.5) -- (3.333333333333333, -2.5);
\draw[thick, red, {[-)}] (3.333333333333333, -2.5) -- (6.666666666666666, -2.5);
\draw[thick, orange, {[-)}] (6.666666666666666, -2.5) -- (7.777777777777777, -2.5);
\draw[thick, blue, {[-)}] (7.777777777777777, -2.5) -- (8.148148148148147, -2.5);
\draw[thick, blue, {[-)}] (8.148148148148147, -2.5) -- (8.518518518518515, -2.5);
\draw[thick, blue, {[-)}] (8.518518518518515, -2.5) -- (8.888888888888886, -2.5);
\draw (8.89, -2.5) node[right]{$2 \times 3^{-1} + 1 \times 3^{-2} + 3 \times 3^{-3}$};

\draw[thick, red, {[-)}] (0, -3.0) -- (3.333333333333333, -3.0);
\draw[thick, red, {[-)}] (3.333333333333333, -3.0) -- (6.666666666666666, -3.0);
\draw[thick, orange, {[-)}] (6.666666666666666, -3.0) -- (7.777777777777777, -3.0);
\draw[thick, blue, {[-)}] (7.777777777777777, -3.0) -- (8.148148148148147, -3.0);
\draw[thick, blue, {[-)}] (8.148148148148147, -3.0) -- (8.518518518518515, -3.0);
\draw[thick, black, {[-)}] (8.518518518518515, -3.0) -- (8.641975308641975, -3.0);
\draw (8.64, -3) node[right]{$2 \times 3^{-1} + 1 \times 3^{-2} + 2 \times 3^{-3} + 1 \times 3^{-4}$};


\draw[thick] (8.6, 0.1) node[above]{0.86} -- (8.6, -3.1);
 
\end{tikzpicture}

\bigskip

Notice the structure of how we would find the $d_i$ in the previous argument: knowing $d_1, \cdots, d_{i-1}$, one would pick $d_i$ as the maximal one such that $\sum_{j = 1}^i d_j b^{k-j} \leq t$. This is, in fact, the exact same algorithm used to find the digits of $t$ in base $b$.

So, let us consider the following algorithm to calculate $g(t)$:
%todo put into words

Let $b^k$ be the least power of $b$ greater than $t$ for $t > 0$.

\begin{lstlisting}
$rem \leftarrow t$
$i \leftarrow 0$
Loop:
  $i \leftarrow i-1$
  Put $d_i$ be the greatest integer $k$ such that $k \cdot b^{k-i} \leq rem$
  $rem \leftarrow rem - d_i b^{k-1}$
\end{lstlisting}
\label{digalg}

The preceding algorithm will compute the digits of $t$ in base $b$. Indeed, in base $b$, $t$ would be represented as:

\[d_1 d_2 \cdots d_k \, . \, d_{k+1} \cdots\]

Or, in scientific notation:

\[b^k \times 0.d_1 d_2 d_3 \dots\]

All the $d_i$ are members of the set $\{ 0, \cdots, b-1\}$, and furthermore we have:

\[\sum_{i = 1}^\infty d_i b^{k-i} = t\]

Finally, for any $N \geq 0$:

\[(\sum_{i=1}^N d_i b^{k-i}) + b^{k-N} \geq t\]

As such, we may consider the following sequence of intervals:

\begin{gather*}
\left\{\, \left[ x , x + b^{k-N} \right[ \;\middle|\; x = \left(\sum_{i=1}^N d_i b^{k-i}\right) + j b^{k-N}, 0 \leq j < d_N, 0 \leq N \leq M\,\right\}\\
\bigcup\\
\left\{\left[ \left(\sum_{i=1}^M d_i b^{k-i}\right) ,\left(\sum_{i=1}^M d_i b^{k-i}\right) + b^{k-M} \right[\right\}
\end{gather*}

To simplify the notation we make the following definitions:

\begin{definition}
First, we define $R_\delta(x)$ to mean $\left[x, x+\delta \right[$.

Let $\{d_i\}_{i \in \N}$ be a sequence of natural numbers. Define the collection of intervals denoted by
\[ b^k [d_0 d_1 d_2 \dots] \]
as follows:

First, define it inductively for finite sequences:

\begin{gather*}
b^k [] = \emptyset\\
b^k [d_0 \cdots d_n d_{n+1}] = b_n [d_0 \cdots d_n] \cup \left\{\, R_{b^{k-(n+1)}}(d_0 . d_1 \cdots d_n j \times b^k) \;\middle|\; 0 \leq j < d_{n+1} \,\right\}.
\end{gather*}

And, for infinite sequences, simply take the increasing union
\[ b^k[d_0 d_1 \dots] = \bigcup_{i = 0}^\infty b^k[d_0 \cdots d_i].\]

\end{definition}

The following proposition is stated without proof.

\begin{prop}
Let $\{d_i\}_{i \in \N}$ be a sequence of natural numbers. Let $A = b^k [d_0 d_1 d_2 \dots]$. Then, $A$ satisfies the following properties:

\begin{itemize}
\item Every interval $I \in A$ is $b$-ary

\item $\cup_{I \in A} I = \left[ 0, t \right[$, where $t = b^k \times d_0 . d_1 d_2 \dots$

\item Every two intervals in $A$ are disjoint

\item The Hausdorff sum of $A$ is simply given by $b^{ks} \sum d_i b^{-is}$.
\end{itemize}

\end{prop}

\begin{remark}
Notice that the Hausdorff sum of $A$ can be easily written out as a $b^s$-ary expansion. Namely, $(b^s)^k \times d_0 . d_1 d_ 2 \dots$.

It is left implied when writing things with scientific notation that the base in which the expansion should be interpreted is the base of the mantissa. In this case, $b^s$. In the future, however, in this particular case, we will abbreviate it to $b^{ks}$, but the expansion following should still be interpreted as base $b^s$.
\end{remark}

Let $t$ have $b$-ary expansion $b^k \times d_0 . d_1 d_2 \dots$. Consider the coverings given by
\[ b^k [d_0 \dots d_{n-1} (d_n + 1)] \]
as $n$ ranges over $\N$.

Any of these covers $t$, and so we have:

\[ g(t) \leq \inf \{\, b^{ks} \times d_0 . d_1 d_2 \cdots d_{n-1} (d_n + 1) \mid n \in \N\,\}\]

This is, in fact, an equality.

\begin{prop} \label{ginfs}
Let $t$ have $b$-ary expansion $b^k \times d_0 . d_1 d_2 \dots$, with $d_0 = 0$.

Define $s_n = b^{ks} \times d_0 . d_1 d_2 \cdots d_{n-1} (d_n + 1)$ for $n \in \N$

Then:

\[ g(t) = \inf s_n\]
\end{prop}

\begin{proof}
We already know $LHS \leq RHS$, so we will focus on showing the opposite inequality.

We will do this in two parts: first, we will show that the infimum required to compute $g(t)$ can be taken over coverings of the form $b^k [a_0 a_1 a_2 \dots]$. Then, we will show that any covering of this form has contribution that is $\geq \inf s_n$.

Fix an arbitrary $b$-ary covering $A = \{ I_i \}_{i \in \N}$ of $\left[ 0, t \right[$.

We may begin by supposing there is no interval of size greater than $b^k$. Indeed, the only intervals greater than this size that are not immediately discardable are those of the form $\left[0, b^m \right[$ with $m > k$, but these can obviously be replaced by $\left[0, b^k \right[$ without any loss of generality. In fact, assume $A$ is wholly contained in $\left[0, b^k \right[$.

Define, inductively, the sequence $\{a_i\}_{i \in \N}$ as follows:

For all $n \in \N$, defined $a_0 \cdots a_{n-1}$, let $a_n$ be the greatest natural such that $b^k [a_0 \cdots a_n] \subseteq \bigcup I_i$. Let $A'$ be the covering given by $b^k [a_0 a_1 \dots]$. We wish to show that this covering not only covers $\left[0, t \right[$, but also has Hausdorff sum less than or equal to that of $A$.

First, we show it covers the desired interval. We know $A'$ covers the interval from 0 up to $a = b^k \times a_0 . a_1 a_2 \dots$, so suppose for the sake of arguent that $a < t = b^k \times d_0 . d_1 d_2 \dots$. It is easy to verify that if $a < t$ then the following happens:

There exists $n$ such that for $i < n$ we have $a_i = d_i$, and $a_n < d_n$.

But if this is the case, we have a contradiction with our definition of $a_n$ by maximality, as it means $b^k \times a_0 . a_1 \cdots a_{n-1} (a_n + 1)$ would be $\leq t$ and so $b^k [a_0 \cdots a_{n-1} (a_n + 1)] \subseteq \bigcup I_i$.

So, we have established that $A'$ covers $\left[0, t\right[$. We will now show its Hausdorff sum is $\leq$ that of $A$.

First, suppose without loss of generality that $A$ does not contain any interval that does not intersect $A'$. Done this, we show that any interval $I$ in $A$ is contained in one of the intervals in $A'$.

Fix, then $I \in A$. Let it intersect $J \in A'$. By nature of the $b$-ary intervals, either $I \subseteq J$ (as desired), or $J \subseteq I$. But, if the latter, we proceed to conclude that $J = I$:

Suppose $I$ is of the form $\left[ i b^{k-n} \oplus b^{k-n} \right[$. By the maximality of $a_n$, we have the following:

\[ \sum_{i = 0}^n a_i b^{k-i} \geq (i+1) b^{k-n} \]

And therefore, $J$ is present in $b^k [a_0 \cdots a_n]$. Hence, the length of $J$ is $\geq b^{k-n}$, and so if $J \subseteq I$ we have $J = I$, as desired.

We can now finish the argument as follows:

For each interval $J \in A'$ collect the intervals $I \in A$ contained in $J$. Note that, by construction, these cover $J$. We will show that each of these collections can be replaced by $J$ without raising the Hausdorff sum, which allows us to conclude, then, that the Hausdorff sum of $A'$ is $\leq$ that of $A$.

By elementary properties of the Lebesgue measure, we have:

\[ \sum \lvert I \rvert \geq \lvert J \rvert \]

From whence follows, by convexity and elementary calculus, that for $s \leq 1$ we have, as desired,

\[ \sum \lvert I \rvert^s \geq \lvert J \rvert^s \]

This concludes the first part of the proof. We move on to the second part.

We have already concluded we need only consider coverings of the form $b^k [a_0 a_1 a_2 \dots]$, and in fact we need only take into account those constructed as above; this will allow us to use certain maximality arguments. We will now show that we may further restrict our attention to those sequences of the form $d_0 d_1 \cdots d_{n-1} (d_n + 1)$.

Fix any arbitrary sequence $\{a_n\}$ constructed as above. We necessarily have $b^k \times a_0 . a_1 a_2 \dots \geq b^k \times d_0 . d_1 d_2 \dots$. Either these two sequences are the same, or there exists a first index $j$ where they differ.

In the latter case, let $n$ be this index. That is, $a_i = d_i$ for $i < n$, $a_n \neq d_n$. It is necessarily the case that $a_n > d_n$, as if $a_n < d_n$ we would reach a contradiction with the maximality of $a_n$. In this case we can safely assume our covering is of the desired form: indeed, the covering given by the sequence $d_0 d_1 \cdots d_{n-1} (d_n + 1) 0 0 0 \dots$ also covers $\left[0, t \right[$, and has Hausdorff sum $\leq$ than that of the covering given by $\{a_n\}$.

As for the former case, consider the sequence $s_n$ we defined as $b^k \times d_0 . d_1 d_2 \cdots d_{n-1} (d_n + 1)$. We remark that $\lim s_n = b^k \times d_0 . d_1 d_2 \dots$. This limit is, in fact, the Hausdorff sum of the covering in consideration (that is, $b^k [d_0 d_1 d_2 \dots]$), and it is elementary to notice that $\lim s_n \geq \inf s_n$, as desired.

We have now, shown, then, that whatever the covering we choose, $\inf s_n$ will be $\leq$ than the Hausdorff sum of our covering, and therefore we have finally shown the result:

\[ g(t) = \inf s_n \]

\iffalse
Now, to show we can assume all intervals to be disjoint: fix any point $a$ in $\cup I_i$. Define $J_a$ to be \emph{the biggest interval in $A$ containing $a$}. We propose, now, that the covering $A' = \{ J_a \}$ covers exactly the same points as $A$, while having a Hausdorff sum less than or equal to that of $A$, as well as having all its intervals be disjoint.

The first two properties are trivial, so we devote our attention only to the third one: suppose, for the sake of argument, that $J_a$ and $J_b$ are not disjoint. The structure of the $b$-ary intervals mandates that either $J_a \subseteq J_b$ or $J_b \subseteq J_a$; suppose the former for the sake of argument. Then, $a \in J_b$, which implies $J_a$ is of length greater than or equal to that of $J_b$ (by definition), and so, in fact, $J_a = J_b$.

This digression completed, we may go back to the problem at hand.
\fi
\end{proof}

This last proposition has a few consequences, one of which is the following: it gives us an algorithm to approximate or even calculate the value of $g(t)$, facilitating its study.

\section{$b$-ary digit expansions}\label{sec7}

We will now use the preceding proposition to look at the function $g$ from another angle. We will begin by taking a closer look at the notion of digit expansion.

\begin{definition}
We define a \emph{digit expansion} as a \emph{bounded} function $d : \Z \to \N$ satisfying the following property:

\begin{center}
There exists $k$ such that for $i < k$ we have $d_i = 0$
\end{center}
\end{definition}

Fixed a base $b \in \left]1, \infty \right[$, to any digit expansion we can assign a real number. Indeed, by definition, the following sum, which we define as the `interpretation of $d$ in base $b$`, converges:

\[ \sum_{i = -\infty}^\infty d_i b^{-i} \]

Recall that any positive real number has a $b$-ary digit expansion. This expansion is not always unique, but the so-called `greedy expansion', the one obtained by the algorithm outlined in page \pageref{digalg}, is. This also happens to be the lexicographically smallest digit expansion whose interpretation is base $b$ is precisely the number we started with. In this expansion, all the digits are contained in the set $\N \cap \left[0, b \right[$. Of course, for natural $b$, this is the familiar set of digits $\{0, \cdots, b-1 \}$, and natural bases have the benefit of almost-uniqueness: indeed, any number is uniquely determined by, and determines uniquely, a decimal expansion, as long as we restrict ourselves to those with digits in the set $\{0, \cdots, b-1 \}$ and that don't end with an infinite sequence of $b-1$.

We will use the notation $\dig_b t$ to mean the greedy digit expansion of $t$ in base $b$. In the other direction, given a digit expansion $d$ we use $\intr_b d$ to mean the interpretation of $d$ in base $b$.

\begin{remark}
It should be noted that there is a slight sloppiness when we say things like ``let $b^k \times d_0 . d_1 d_2 \dots$ be the $b$-ary expansion of $t$''. Indeed, what we are saying here is not that the sequence $d$ is the expansion of $t$, at least not in the sense above. Rather, the actual expansion would be something defined as follows:

\[d'_i =
\begin{cases}
d_{i-k} & \text{for $i \geq k$}\\
0 & \text{for $i < k$}
\end{cases}
\]

Furthermore, we often use that same notation both to refer to extracting the digits from a number, as well as to interpret a sequence of digits. Hopefully, however, this should not cause any confusion.
\end{remark}

We now propose an alternate definition of the function $g$.

As always, suppose fixed a base $b$ and a dimension $s$. We begin by defining, for $t \geq 0$:

\[ i_{bs}(t) = \intr_{b^s} \dig_b t \]

This function will appear regularly in the sequence.

Now, we claim the following:

\begin{theorem} \label{ginfi}
\[g_{bs}(t) = \inf_{x \geq t} i_{bs}(t)\]
\end{theorem}

\begin{proof}
Suppose $t$ has $b$-ary expansion $b^k \times d_0 . d_1 d_2 \dots$.

We have already established $g(t)$ can be calculated as $\inf s_n$, where $s_n = b^{sk} \times d_0 . d_1 d_2 \cdots d_{n-1} (d_n + 1)$. It is easy to notice that we can restrict the $n$ to those such that $d_n \neq b-1$, and therefore those that are determined by their interpretation in base $b$. That is,

\[g(t) = \inf_{n \mid d_n \neq b-1} i(t_n) \]

Where $t_n = b^k \times d_0 . d_1 d_2 \cdots d_{n-1} (d_n + 1)$.

This, then, shows $LHS \geq RHS$.

For the other inequality, fix any $x \geq t$. We already know that $i(t) \geq g(t)$, so we can assume $x > t$.

We can, furthermore, assume $x < b^{k+1}$. Indeed, if $x \geq b^{k+1}$ we would have $i(x) \geq b^{(k+1)s}$, which is greater than $s_0 = b^{ks}$, and so any $x \geq b^{k+1}$ can be safely disregarded.

We restrict our attention to $x$ between $t$ and $b^{k+1}$, then, and so we may assume $x$ has a $b$-ary expansion $b^k \times x_0 . x_1 x_2 \dots$. We will now show we can restrict our attention to those of the form $b^k \times d_0 . d_1 d_2 \cdots d_{n-1} (d_n + 1)$.

Fix such an arbitrary $x > t$. Let $n$ be the index such that $d_i = x_i$ for $i < n$ and $d_i \neq x_i$. Then, it must be the case that $d_n < x_n$, and in that case we simply point out that the number $t_n$ given by $b^k \times d_0 . d_1 d_2 \cdots d_{n-1} (d_n + 1)$ (whose $b$-ary expansion really is this one because it must be the case that $d_n + 1 \leq b-1$) satisfies $i(t_n) \leq i(x)$, which concludes our proof.
\end{proof}

\section{$g$ as a function of $s$}

We now present an application of theorem \ref{ginfi} which will be useful in the sequence.

\begin{prop} \label{gfuncofs}
Fix arbitrary $t > 0$ and a base $b$. Define $h(s) = g_{bs}(t)$.

Then, $h$ is a continuous function from $\left]0, 1 \right]$ to $\R$. Furthermore, for $t \in \left[0, 1 \right]$, $h$ is decreasing.
\end{prop}

\begin{proof}
We begin by showing $h(s)$ is decreasing. Fix dimensions $s \leq s'$.

Notice that for any $t \in \left[0, 1 \right]$ the function $t^s$ is a decreasing function. We will show, based on this, that $i_{bs}(t)$ is decreasing (as a function of $s$), and this clearly shows, since $g_{bs}(t) = \inf_{x \geq t} i_{bs}(t)$, that $g$ is decreasing as a function of $s$.

But indeed, this is a triviality, as if $t \in \left[0, 1\right]$ then it can be expressible with a $b$-ary expansion of the sort $d_0 . d_1 d_2 \dots$, and so $i_{bs}(t) = \sum d_i b^{-si}$, a sum of functions of the form $c t^s$ with $t < 1$. Each of these are decreasing, and so their sum is too, which concludes the first half of the proof.

For the second half, we remind the reader that $g$ can be defined as

\[g(t) = \inf s_n(t)\]

Where, if we let $t$ have $b$-ary expansion $b^k \times d_0 . d_1 d_2 \dots$, $s_n(t)$ is defined as $b^{ks} \times d_0 . d_1 \cdots d_{n-1} (d_n + 1)$.

To show $g_{bs}(t)$ is a continuous function of $s$, it is not quite enough to show all $s_n(t)$ are continuous functions of $s$. It is, however, enough to show they are all Lipschitz with respect to a fixed constant $c$.

Indeed, notice that for any $n$ and $t$ the function $s_n(t)$ (of $s$) is a finite sum of terms of the form $\sum k_i b^{-is}$.  We can, then, take the derivative with regard to $s$, getting derivatives of the form $\sum - i k_i \log b \cdot b^{-is}$. Since all $k_i$ are $\leq b$, this sum is, in absolute, less than $b \log b \sum i b^{-is}$, which converges to $\frac{b^{s+1} \log b}{(b^s + 1)^2}$ for $s > 0$.

Now the argument can be easily completed: fix any $s > 0$. These derivatives are, in the interval $\left[s, 1 \right]$, all bound by $c = \frac{b^{s+1} \log b}{(b^s + 1)^2}$, and so, in this interval, every $s_n(t)$ is $c$-Lipschitz as a function of $s$, and therefore so is $h$. This concludes the proof.
\end{proof}

\section{On notation}

Now that the main ingredients are in place, we can give names to the objects of studies to help us refer to them.

Our main object of study is the `optimal constant' $C$ for the inequality $\BB_{bs} \leq C \HH_{bs}$. This constant depends on $b$ and $s$, and so, fixed such a base and dimension, we call this optimal constant $C_{bs}$.

We have already shown an upper bound to this constant is given by the solution to the optimization problem:

\[ \sup_{\substack{\ell,r \in \R^+_0\\\ell+r \neq 0}} \frac{g_{bs}(\ell) + g_{bs}(r)}{(\ell + r)^s} \]

We will call the solution to this optimization problem $K_{bs}$.

Subscripts will be omitted when unnecessary when fixed or arbitrary $b$ and $s$ are implied.

\section{On calculating $K$} \label{calck}
%This section could go before introducing the bases, maybe?

In order to solve our optimization problem, we first make an observation about the function $g$: it is self-similar. Indeed, $g(b t) = b^s g(t)$. As such, it grows like $t^s$.

To be more rigorous, consider the interval $\left[ 1/b, 1 \right]$. The function that takes $t$ into $g(t)/t^s$ is continuous in this interval, and so, by Weierstrass, has a maximum and a minimum.

Indeed, by convexity, $i(t) \geq t^s$, which, by the definition of $g$ using $i$, makes the inequality $g(t) \geq t^s$ evident. Equality is attained at, for example, $t = 1$.

The maximum turns out to be much harder to pinpoint, but in fact it turns out to be very closely related to $K$. To see why, let us give it a name, $\Theta_{bs}$, and let, for the moment, $\theta$ be a maximizer. It is clear that, while we took the maximum in $\left[ 1/b, 1 \right]$, the inequality $g(t) \leq \Theta t^s$ applies over all $\R^+_0$. As such, we get:

\[\frac{g_{bs}(\ell) + g_{bs}(r)}{(\ell + r)^s} \leq \frac{\Theta \ell^s + \Theta r^s}{(\ell+r)^s}\]

The right hand side can be easily maximized with elementary calculus, giving us a maximum at $\ell = r$ of $2^{1-s} \Theta$. As such, we get one inequality:

\[ 2^{1-s} \Theta \geq K \]

The other inequality can be shown by noticing that, for $\ell = r = \theta$ the quantity to be maximized amounts to precisely $2^{1-s} \Theta$. This shows, then, that to calculate an upper bound of $C$ we can focus on trying to bound $g$ from above as closely as possible by a function of the form $\Theta t^s$. In other words, we are now interested in the optimization problem:

\[ \max_{t \in \left[ 1/b, 1 \right]} \frac{g(t)}{t^s} \]

Notice that the function being optimized is a quotient of two increasing functions. There exists an algorithm to approximate, to arbitrary precision, this class of functions. We present a naïve algorithm to this effect.
%note: i have a less naïve algorithm. should i present that?

\section{Optimization of increasing function quotients} \label{optimization}

Suppose fixed an interval $I = \left[ a, b \right]$ and two increasing functions $I \to \R^+$, call then $u$ and $v$. We focus on approximating the solution to the optimization problem:

\[ \max_{t \in I} \frac{u(t)}{v(t)} \]

For purposes of argument, let $S$ be the maximum we wish to find.

Fix a partition $P$ given by $a = t_0 < t_1 < \cdots < t_n = b$ of $I$. For $i = 0, 1, \cdots, n-1$ define $M_i = \frac{u(t_{i+1})}{v(t_i)}$ and $m_i = \frac{u(t_i)}{v(t_{i+1})}$. Clearly, $M_i$ and $m_i$ are upper and lower bounds, respectively, of $u/v$ in the interval $\left[t_i, t_{i+1} \right]$. As such, if we put $M(P) = \max M_i$ and $m(P) = \max m_i$, we have the inequality:
%meanwhile i realized the choice for m_i is dumb. should i mention that?
%would be more efficient to just do u(t_i)/v(t_i), as thats a lower bound of the max, and a slightly tighter one at that

\[m(P) \leq S \leq M(P)\]

Of course, as is usual with this type of argument, refining the partition will get us tighter bounds. Indeed, if $P_n$ is a sequence of partitions whose diameter converges to zero, $M(P_n) - m(P_n) \to 0$, and so each of these sequences will converge to $S$. This concludes this section on this kind of optimization problem.

\section{On the continuity of $K$}

In this section, we present the following result:

\begin{theorem}
As a function of $s \in \left]0, 1\right]$, the quantity $K_{bs}$ is continuous.
\end{theorem}

To this effect, we use the result of section \ref{calck} to reduce the matter to showing the following proposition:

\begin{prop}
As a function of $s \in \left]0, 1\right]$, the quantity $\Theta_{bs}$ is continuous.
\end{prop}

Notice that we defined $\Theta$ as follows:

\[ \Theta_{bs} = \max_{t \in \left[ 1/b, 1 \right]} \frac{g(t)}{t^s} \]

As such, assuming $g_{bs}(t)$ is continuous as a function of $(s, t) \in \left]0, 1\right] \times \left[1/b, 1\right]$, which we will do shortly, we need only prove the following lemma:

\begin{lemma}
Let $X$ and $J$ be subsets of $\R$, and suppose $J$ is compact. Let $f$ be a continuous function $X \times J \to \R$.

We can define a function $u(x) : X \to \R$ defined as:

\[ u(x) = \max_y f(x,y) \]

This function is continuous.
\end{lemma}

\begin{proof}
We prove this lemma by supposing fixed arbitrary $x_0 \in X$ and a sequence of $x_n \in X$ converging to $x_0$. We will show $u(x_n)$ converges to $u(x_0)$.

Fix arbitrary $\varepsilon > 0$, and let $y_0$ be a maximizer of $f(x_0, y)$. There exists a neighbourhood, let's say $B_{\delta_1}(x_0) \times B_{\delta_2}(y_0)$ such that, for $(x,y)$ in this neighbourhood, $f(x,y)$ is within $\varepsilon$ of $f(x_0, y_0) = u(x_0)$. As such, for $x \in B_{\delta_1}(x_0)$ we have $u(x) \geq u(x_0) - \varepsilon$.

On the other hand, consider the succession $u(x_n)$. We can, without loss of generality, pass to a converging subsequence with limit $L$. The preceding argument shows $L \geq u(x_0)$; we now show the reverse inequality.

Let $y_n$ be a sequence of maximizers, that is, suppose it satisfies $f(x_n, y_n) = u(x_n)$. Suppose, again without loss of generality, that $y_n$ converges, to a limit $y_L$.

It is a fact that $L = \lim f(x_n, y_n) = f(x_0, y_L)$ by continuity. But, furthermore, $f(x_0, y_L) \leq u(x_0)$, which finishes our argument.
\end{proof}

To conclude the section, we tie the following loose end:

\begin{prop}
Let $f : \left]0, 1\right] \times \left[1/b, 1\right] \to \R$ be defined as

\[f(s, t) = g_{bs}(t)\]

Then, $f$ is continuous.
\end{prop}

\begin{proof}
Notice that, fixed $s$, $f$ is a continuous increasing function of $t$. Furthermore, as per proposition \ref{gfuncofs}, it is also, fixed $t \in \left[1/b, 1 \right]$, a continuous decreasing function of $s$. Elementary arguments show that, under these hypotheses, $f$ is continuous. %maybe i should be a bit less lazy here, huh?
\end{proof}

\section{A lower bound for $C$}\label{seclowerbound}

In the previous sections, we have constructed an upper bound for $C$. Namely, $C \leq K$. We do not, however, have any nontrivial lower bounds yet. We seek to remedy that.

First, a few definitions.

Notice that $g$ is a continuous increasing function, surjective from $\R^+_0$ to $\R^+_0$. This means that, for any $y$, there is a least $x$ such that $g(x) = y$. We will denote this by $x = \len_{bs}(y)$.

A quantity that will be useful to us is $\len_{bs}(1)$, which we will denote by $\gamma_{bs}$. It is obvious that $\gamma \in \left] 1/b, 1 \right]$.

This quantity is useful because of the following: consider the interval $\left[0, \gamma \right[$. There are many ways to cover it using binary intervals, but we can, without loss of generality, assume it is covered by the single interval $\left[0, 1 \right[$. Of course, this holds up to scaling by powers of $b$, which allows us to make certain constructions. In particular, we will use it to make a construction with the intent of maximizing the discrepancy between the Hausdorff and $b$-ary measure.

\textbf{In what follows, we assume tacitly $0 < s < 1$.}

Fix an arbitrary dimension $s \in \left]0, 1 \right[$ and base $b$. Build, inductively, $S_n$ as follows:

Put $S_0 = \left]-\gamma, \gamma\right[$.

Now, to construct $S_1$, we begin by looking at $S_0$ as a countable union of $b$-ary intervals. Divide each of these into $b^\alpha$ equal intervals ($\alpha \geq 0$). Then apply the following transformation to each of them:

If $I = \left[a \oplus b^\eta \right[$ is a $b$-ary interval, its transform is given by:

\[\left] c - b^{\eta - \beta} \gamma, c + b^{\eta - \beta} \gamma \right[\]

Where $c = a + b^{\eta - 1}$.

Notice that this interval can be decomposed into countably many disjoint $b$-ary intervals, as per the $b$-ary expansion of $\gamma$.

An important property of this transformation is as follows: \emph{The smallest $b$-ary interval covering the entire transform of $I$ is $I$ itself.}
%todo show that gamma is nonterminating

Now, inductively, suppose $S_n$ is a countable disjoint union of $b$-ary intervals. To each of these, do the same as  we did to $S_0$: divide it into $b^\alpha$ parts, and transform each of these parts as indicated above. This will net us $S_{n+1}$.

This process will give us a decreasing sequence of sets, whose intersection we call $S$. Notice that $S$ depends on a few factors we left implicit: namely, $\alpha$, $\beta$, $b$ and $s$.

We wish to manipulate these parameters to maximize the difference between the Hausdorff and $b$-ary coverings of $S$. We begin by inspecting the Hausdorff measure.

Note that all of the $S_i$, taken as unions of intervals, are Hausdorff covers of $S$. Let $\sigma_i$ be the Hausdorff sum of $S_i$.

Clearly, $\sigma_0 = (2 \gamma)^s$. Furthermore, by the recursive definition of $S_n$, $\sigma_n$ is a geometric series, whose ratio between successive terms is simply $\frac{\sigma_1}{(2 \gamma)^s}$.

The only way for the limit to be nontrivial is for this series to be the constant one, and so for $\sigma_1$ to be $(2 \gamma)^s$. As such, let us calculate $\sigma_1$.

First, calculate the Hausdorff sum of what the transformation would do to $\left[0, 1 \right[$. It would be turned into $\left] \frac 1 b - \frac \gamma{b^\beta}, \frac 1 b + \frac \gamma{b^\beta} \right[$, with Hausdorff sum $(2 \gamma)^s b^{-s \beta}$. Now, if this transformation was applied to an interval of length $\ell$, then by rescaling the result would have Hausdorff sum $(2 \gamma)^s b^{-s \beta} \ell^s$.

Now, notice that we are not merely taking intervals and applying the transformation to them. First, we split the interval into $b^\alpha$ many. Hence, from an interval of length $\ell$, the result actually has Hausdorff sum equal to $b^\alpha (2 \gamma)^s b^{-s (\alpha + \beta)} \ell^s$.

If we have a set of intervals with Hausdorff sum $H$ and apply this to all of them, the resulting Hausdorff sum is given by:

\begin{align*}
\sum_\ell (2 \gamma)^s \ell^s b^{\alpha - (\alpha + \beta) s} &=
(2 \gamma)^s b^{\alpha - (\alpha + \beta) s} \sum \ell^s\\
&= (2 \gamma)^s b^{\alpha - (\alpha + \beta) s} \cdot H
\end{align*}

But we are not doing something as direct as that. First, we are splitting our intervals into $b$-ary ones. The effect of this is it tampers with the value of $H$.

All the intervals created in this construction are of the form $\left] c - b^\eta \gamma, c + b^\eta \gamma \right[$. As such, splitting it into $b$-ary intervals yields a collection of intervals with Hausdorff sum equal to $2 b^{\eta s}$. Hence, splitting this into $b$-ary intervals and applying the above to it, we get a final result of:

\[\sigma_1 = (2 \gamma)^s b^{\alpha - (\alpha + \beta) s} \cdot 2 \]

We need this to be equal to $\sigma_0 = (2 \gamma)^s$, and so we meddle with the parameters in order to make this true. In particular, we solve the equation:

\[ 2 b^{\alpha - (\alpha + \beta) s} = 1 \]

Yielding the result

\[s = \frac{\alpha + \log_b 2}{\alpha + \beta}\]

For this particular dimension, we have shown $S$ has Hausdorff measure \emph{at most} $(2 \gamma)^s$. We will now bound from below the binary measure.

First, a technical lemma:

\begin{lemma}
Let $S \subseteq \R$ be compact. Then, the $s$-dimensional $b$-ary measure of $S$ can be calculated through finite $b$-ary coverings of $S$.
\end{lemma}

\begin{proof}
Clearly, any finite $b$-ary covering is a $b$-ary covering. We need only show that any $b$-ary covering can be arbitrarily approximated by finite ones.

To this effect, fix a covering $\{ I_i \}_{i \in \N}$, where $I_i = \left[x_i, y_i \right[$. Augment it with the $b$-ary intervals $J_i = \left[x_i - \varepsilon_i, x_i \right[$, where $\varepsilon_i$ are such that their sum is at most some arbitrary $\varepsilon$. Then, the Hausdorff sum of the collection formed by all $I_i$ and $J_i$ is arbitrarily close to that of the initial covering.

Now, consider the collection of intervals $\{ \left]x_i - \varepsilon_i, y_i \right[ \}_{i \in \N}$. This is an open cover of $S$, and so, a finite subcovering, say $\{ \left]x_i - \varepsilon_i, y_i \right[ \}_{i \in \mathcal{I}}$, covers $S$. As such, we consider the collection of intervals of $I_i$ and $J_i$ with $i \in \mathcal{I}$; this collection is finite, covers $S$ and its Hausdorff sum is at most that of the original one plus an arbitrary $\varepsilon$, concluding our proof.
\end{proof}

This lemma is useful because $S$ is compact.

Indeed, it is trivial to see that the closure of $S_{n+1}$ is contained in $S_n$. As such, $S$ can be written as the intersection of all $\overline{S_n}$, all of which are compact, hence $S$ itself is compact. Therefore, we may bound the $b$-ary measure of $S$ from below by minoring the Hausdorff sum of \emph{finite} $b$-ary coverings of $S$.

We can actually go a step further, in fact. Notice that, for $n > 0$, $S_n$ contains no points of the form $k b^{-n}$. This implies that, fixed a finite $b$-ary covering of $S$, it actually covers some $S_n$.

To show this is so, fix any finite $b$-ary covering of $S$. Let $n$ be a natural number such that all intervals of this covering have length greater than $b^{-n}$. Consider $S_n$ as a countable collection of intervals.

Notice that each of these intervals is either wholly contained in the covering we have fixed, or entirely outside it. Of course, if any of them were outside the covering, there would be at least one point of $S$ not being covered, which is a contradiction. Hence, $S_n$ is entirely covered.

Therefore, we now work with the following hypothesis: \textbf{Fix a \emph{finite} covering of some $S_n$}. If we can bound from below the Hausdorff sum of coverings of this kind, we have bound from below the $b$-ary measure of $S$, as desired.

We will now show by induction on $n$ that this covering has Hasudorff sum at least two.

If $n = 0$ the result is a mere consequence of the definition of $\gamma$. We will now suppose the result to be true for some $n$, and show it for $n+1$.

Focus on the part of $S_{n+1}$ that is to the right of the origin. Let $0.d_1 d_2 \dots$ be the digits of $\gamma$, and let $\{I_i\}$ be the intervals of $b^0 [0 d_1 d_2 \dots]$. Within each $I_i$ are $b^\alpha$ copies of $S_n$, scaled down by a factor of $\text{[Length of $I_i$]} \times b^{-(\alpha + \beta + 1)}$. If we show we can assume any covering of $S_{n+1} \cap I_i$ covers the entirety of $I_i$, then we may suppose any covering of $S_{n+1}$ covers the entirety of $\left]-\gamma, \gamma\right[ = S_0$, concluding the argument.

The only missing piece of the puzzle, then, is to show that we can suppose that $I_i$ is completely covered. To simplify the argument, we assume $I_i$ to be the unit interval; it is a mere matter of scaling it down to make it work for arbitary $b$-ary intervals.

There are two cases: either we cover the whole of our interval $I$ (which is the desired case) or we have a separate cover for each of its $b$ sub-intervals. By symmetry, if $\alpha \geq 1$, we may assume all of these covers to be the same. Repeating the argument (if $\alpha \geq 2$), we can assume we have equal covers for the $b^2$ sub-intervals of $I$. Repeat $\alpha$ times to get that, either we cover the entire interval, or we find some covering of one of the scaled down copies of $S_n$. There are $b^\alpha$ such copies; if we show any covering of one of them has Hausdorff sum at least $b^{-\alpha}$ the argument is complete.

Consider such a scaled down copy, let $c$ be its center, and fix an arbitrary $b$-ary covering of it. If this covering has an interval that crosses over $c$, this interval must have length at least $b^{-\alpha}$, and so its Hausdorff sum is at least $b^{- \alpha s} \geq b^{-\alpha}$. We may then suppose our covering does not pass through the center. But then, by rescaling and translation, we reduce the problem to finding a covering of $S_n$. We already know any such covering has Hausdorff sum at least two, and so any covering of our scaled down copy has Hausdorff sum at least

\[ 2 b^{-(\alpha + \beta) s} = b^{-\alpha} \]

Which concludes our proof. \qed

To recap: our set $S$, built for any dimension of the form $s = \frac{\alpha + \log_b 2}{\alpha + \beta}$, has:

\begin{gather*}
\HH_s(S) \leq (2\gamma)^s\\
\BB_s(S) \geq 2
\end{gather*}

Which shows that, for these dimensions,

\[C \geq 2^{1-s} \gamma^{-s}\]

We have, then, a partial bound. If we define $\Gamma_{bs} = \gamma_{bs}^{-s}$, we have the partial bounds:

\[ 2^{1-s} \Gamma_{bs} \leq C_{bs} \leq 2^{1-s} \Theta_{bs} \]

Where we have seen the right inequality always holds, and the left equality holds in a dense subset of $\left[0, 1 \right]$.

\section{Application}\label{secfinal}

We use the results from the preceding section to construct a set $S$ satisfying the following property:

\[ \BB_{5s}(S) < \BB_{4s}(S) \]

For some dimension $s$.

This is useful because it disproves the conjecture that, as $b$ increases, so does $\BB_{bs}$.

In view of what we have already done, the set $S$ is not difficult to construct. Indeed, simply replicate the construction done above for $b = 4$, $\alpha = 0$  and $\beta = 1$, getting $s = 1/2$.

For this $S$ we already know $\BB_{4s}(S) \geq 2^{1-s} \Gamma_{4s} \HH_s(S)$.

We know, furthermore, that $\BB_{5s}(S) \leq K_{5s} \HH_s(S) = 2^{1-s} \Theta_{5s} \HH_s(S)$. We need only show, now, that  $\Theta_{5s} \leq \Gamma_{4s}$.

We begin by calculating $\gamma_{4s}$. We assert it is given by $4^0 \times 0.111\dots$, which amounts to $1/3$.

To show this, we begin by noticing that for any $t > 1/3$ we have $i(t) \geq 1$, which, joined with the fact that $i(1/3) = 1$ shows that $g(1/3) = 1$.

Indeed, if $1 > t > 1/3$ (the case $t \geq 1$ is trivial), its ternary digits will agree with those of $1/2$ until some $n$-th digit, which will be a $2$ or a $3$. We have, then,

\begin{align*}
i(t) &\geq \frac 1 {4^s} + \frac 1 {4^{2s}} + \cdots + \frac 1 {4^{(n-1)s}} + \frac 2 {4^{ns}} \\
&= \frac 1 {2} + \frac 1 {2^2} + \cdots + \frac 1 {2^{n-1}} + \frac 2 {2^n} \\
&= 1
\end{align*}

Now, notice that the numbers given by $t_n = \frac 1 2 - 4^{-n}$ converge to $1/2$, and $i(t_n)$ is always $< 1$, implying that for any $t < 1/3$ we have $g(t) < 1$.

We conclude, then, $\gamma_{4s} = 1/3$, whence $\Gamma_{4s} = \sqrt 3$.

We need now bound $\Theta_{5s}$ from above. To this effect, we use the strategies outlined in section \ref{optimization}

Consider the following partition of $\left[1/5, 1 \right]$:

\emph{Digit expansions should from now on be assumed to be in base 5}

\begin{center}
\begin{tabular}{| c | @{\qquad\qquad} | c | @{\qquad\qquad} | c | @{\qquad\qquad} | c |}
\hline
0.1 & 0.11 & 0.12 & 1 \\
\hline
\end{tabular}
\end{center}

Using the methods of section \ref{optimization}, and using proposition \ref{ginfs} to calculate $g(t)$, we construct the following table:

\begin{center}
\tabulinesep=1.2mm
\begin{tabu}{| c | | c | @{\qquad\qquad} | c | @{\qquad\qquad} | c | @{\qquad\qquad} | c |}
\hline
& 0.1 & 0.11 & 0.12 & 1 \\
\hline
$g(t)$ & & $\frac 1 {\sqrt{5}} + \frac 1 5$ & $\frac 1 {\sqrt{5}} + \frac 2 5$ & $1$ \\
$t^{-s}$ & $\sqrt 5$ & $\frac 5 {\sqrt 6}$ & $\frac 5 {\sqrt 7}$ & \\
\hline
\end{tabu}
\end{center}

And we then conclude that $\Theta_{5s}$ is bounded from above by

\[ \max \{ (\frac 1 {\sqrt{5}} + \frac 1 5) \sqrt 5, (\frac 1 {\sqrt{5}} + \frac 2 5) \frac 5 {\sqrt 6}, \frac 5 {\sqrt 7} \}\]

Since all of these are strictly less than $\sqrt 3$, we conclude the desired result. \qed


\begin{thebibliography}{1}
\bibitem{falconer}
K. J. Falconer, \textit{The Geometry of Fractal Sets}

\bibitem{rogers}
C. A. Rogers, \textit{Hausdorff Measures}

\bibitem{brownian}
Shah, Jay. \textit{Hausdorff dimension and its applications.}

\bibitem{fefferman}
Fefferman, Charles. \textit{The multiplier problem for the ball.}

\end{thebibliography}


\end{document}
