\documentclass[addpoints]{exam}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{mathtools}

\usepackage{float}
\usepackage{tikz}

\DeclarePairedDelimiter{\Norm}\lVert\rVert
\newcommand{\R}{\mathbb{R}}
\newcommand{\cm}{\mathrm{cm}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\dl}{\mathrm{d}}
\newcommand{\diff}[1]{\frac{\dl}{\dl #1}}

\pagestyle{headandfoot}
\runningheader{MATH 152 Section 45}{Final Exam Prototype}{Duarte Maia}
\runningheadrule
\footer{}{\ifincomplete{Exercise continues on next page.}{}}{}


\qformat{\textbf{Exercise \thequestion.} (Total points: \totalpoints) \hfill}
\renewcommand{\partlabel}{Part (\thepartno)}


\begin{document}

\begin{center}
\Large Final Exam Prototype

MATH 152 - Section 45

\smallskip

\large Section Instructor: Duarte Maia

\smallskip

No date.
\end{center}

\begin{center}
\fbox{
\parbox{6in}{
\begin{itemize}
\item You have 2 hours to answer the questions below on the provided examination book.

\item Write your name and UCID in the cover page of the examination book.

\item You may ask for additional paper. If you do, sign both exam books and number them.

\item Unless otherwise specified, you do not need to justify your answers. However, you must still show your work. You may lose points if I deem you to have performed a significant leap in logic.

\item This exam is graded out of 100 points.
\end{itemize}
}
}
\end{center}

\noindent Useful Formulas:
\begin{itemize}
\item Half-angle formulas:

\begin{tabular}{c c}
$\cos^2(x) = \frac12 + \frac12 \cos(2x)$ &
$\sin^2(x) = \frac12 - \frac12 \cos(2x)$
\end{tabular}
\item Angle sum formulas:

\begin{tabular}{c c}
$\cos(x+y) = \cos(x)\cos(y) - \sin(x)\sin(y)$\\
$\sin(x+y) = \sin(x)\cos(y) + \cos(x)\sin(y)$
\end{tabular}
\item Some derivatives:

\begin{tabular}{c c}
$\diff x \arcsin(x) = \frac1{\sqrt{1-x^2}}$ &
$\diff x \arctan(x) = \frac1{1+x^2}$
\end{tabular}
\end{itemize}

\pagebreak

\begin{questions}
\question
\begin{parts}
\part[5] riemann sum comp
\part[5] riemann sum concept
\end{parts}

\question
\begin{parts}
\part[10] shape area xy
\part[5] initial value problem
\part[5] study func def via int
\end{parts}

\question
\begin{parts}
\part[5] which funcs are invtible
\part[5] study inv
\end{parts}

\question
\begin{parts}
\part[5] exp model
\part[5] diff bases
\end{parts}

\question
\begin{parts}
\part[5] int
\part[5] int
\part[5] int
\part[5] int
\part[5] int
\part[5] int
\part[5] tricky int
\end{parts}

\question[5]
first order ode

\question
\begin{parts}
\part[5] concept
\part[5] concept
\end{parts}
\end{questions}
\end{document}