% !TeX root = main.tex
\section{The Maslov Index}

\subsection{The General Definition}

Let $H$ be a periodic Hamiltonian with flow $\phi_t$. The Maslov index is an integer associated to a nondegenerate contractible $T$-periodic orbit $\gamma$. The process is involved, but in summary:
\begin{enumerate}[algorithm]\label{page:maslovalg}
\item\label{maslovalg:step1} Since $\gamma$ is periodic, it may be seen as a smooth map $S^1 \to M$.
\item Pick an embedding of the disk $D^2 \to M$ whose restriction to the border coincides with $\gamma$.
\item Pick smooth functions $Z_1(x), \dots, Z_{2n}(x) \colon D^2 \to TM$, such that $\{Z_i(x)\}$ forms a symplectic basis for $T_x M$.
\item Now that $T_{\gamma(t)} M$ has a chosen basis for each $t$, we may write the derivative of the flow at $x$, i.e. the map $(\dl \phi_t)_x$, as a $2n \times 2n$ symplectic matrix, which we will call $A(t)$. Note that $A(t)$ will not be periodic.
\item\label{maslovalg:step5} Given a path of symplectic matrices, such as $A(t)$, where $A(T)$ has no eigenvalue equal to one, it is possible to associate to it an integer called the Conley-Zehnder index.
\end{enumerate}

In turn, it is now necessary to compute the Conley-Zehnder index of a path of matrices $A(t)$. To do so first requires one to establish the existence of a map from the symplectic matrices to $S^1$.

\begin{theorem}\label{rhodef}
There exists a unique map $\rho \colon Sp(2n) \to S^1$ which satisfies the following properties.
\begin{enumerate}[label=\roman*)]
\item If $A$ and $T$ are symplectic matrices, $\rho(T A T^{-1}) = \rho(A)$,
\item If $A$ and $B$ are symplectic matrices,
\begin{equation}
\rho\left(\begin{bmatrix} A & 0 \\ 0 & B \end{bmatrix}\right) = \rho(A) \rho(B),
\end{equation}
\item\label{rhodef:xy} If $A$ is a symplectic matrix of the form $A = \begin{bmatrix} X & -Y \\ Y & X \end{bmatrix}$ then
\begin{equation}
\rho(A) = \det(X + \I Y).
\end{equation}
\item\label{rhodef:realev} If all eigenvalues of $A$ are real, then
\begin{equation}
\rho(A) = (-1)^{m_0/2},
\end{equation}
where $m_0$ is the number of negative eigenvalues, counted with multiplicity,
\item $\rho(A^\transposed) = \rho(A^{-1}) = \conj{\rho(A)}$.
\end{enumerate}

Furthermore, if $A$ is a symplectic matrix whose eigenvalues are all distinct, $\rho(A)$ may be computed using the following formula
\begin{equation}\label{rhoformula}
\rho(A) = (-1)^{m_0/2} \prod \lambda^{\sign \Im \omega(\conj v, v)},
\end{equation}
where the product is taken over the eigenvalues of $A$ with positive imaginary part and unit norm, and to each such $\lambda$ we have associated  a (in general, complex) eigenvector $v$, and we have implicitly extended the 2-form $\omega$ on $\R^{2n}$ in the obvious way to $\C^{2n}$.
\end{theorem}

\begin{proof}
See \cite{audin}, Theorem 7.1.3 on page 192.
\end{proof}

We are now ready to define the Maslov index of a path of symplectic matrices $A(t)$:

\begin{enumerate}[algorithm]
\item Define $Sp(2n)^*$ as the collection of symplectic matrices that do not have 1 as an eigenvalue. It is a known fact \cite[proposition~7.1.4]{audin} that $Sp(2n)^*$ is composed of two path-connected components, $Sp(2n)^+$ and $Sp(2n)^-$, labeled by the sign of $\det(A-I)$. Note that $A(T) \in Sp(2n)^*$, so it must lie in one of these two components.
\item\label{maslov:step2} Define the matrices $W^+$ and $W^-$ as
\begin{equation}
W^+ = - I, \quad W^- \mleft[\begin{array}{c|c}
\begin{matrix} 2 & 0 \\ 0 & 1/2 \end{matrix} & 0\\
\hline
0 & -I
\end{array}\mright].
\end{equation}
Note that $W^\pm \in Sp(2n)^\pm$, and so any matrix in $Sp(2n)^*$ can be connected via a path (in $Sp(2n)^*$) to either $W^+$ or $W^-$. Consequently, we extend $A(t)$ to the interval $\interval 0 {2T}$, where for $t \in \interval T {2T}$ the path $A(t)$ represents such a path connecting $A(2T)$ to $W^\pm$.
\item Define $\alpha(t) = \rho(A(t))$ for $a \in \interval 0 {2T}$. It is trivial to check using property \ref{rhodef:realev} from theorem \ref{rhodef} that $\rho(W^\pm) = (-1)^n$, and also that $\rho(I) = 1$. Therefore, $\rho \circ \alpha$ is a path in $S^1$ starting at $1$ and ending at $1$ or $-1$. Therefore, taking the square (viewing $S^1 \subseteq \C$) one obtains a loop $\rho^2 \circ \alpha$ starting and ending at $1$.
\item Now that we have a loop in $S^1$, we can obtain an integer by considering its winding number: the number of times the loop does a full turn around the circle. It is necessary to establish which direction is positive. Generally, one considers the counterclockwise direction to be the positive direction, but for purposes of the Conley-Zehnder index we count the number of \emph{clockwise turns}.
\end{enumerate}
The integer resulting from the above procedure is called the Conley-Zehnder index of the path $A(t)$, denoted $\mu(A(t))$, and finally we define the Maslov index of the periodic orbit $\gamma$ above, also denoted with the symbol $\mu(\gamma)$, as the Conley-Zehnder index of the corresponding matrix path $A$.

It is a very nontrivial fact that the corresponding integer does not depend on any of the choices done in the process, only on the path and on the Hamiltonian flow. The details may be found in \cite{audin}, whose whole 7th chapter is dedicated to the definition, and well-definition, of this index.

\subsection{The Particular Case of $Sp(2)$}\label{sec:maslovsp2}

The theory of the Maslov index simplifies greatly in the two-dimensional case. This is partly due to the much greater control over the spectrum of a matrix. In two dimensions, the symplectic matrices coincide with those whose determinant equals one, and so their spectrum is entirely determined by their trace.

\begin{prop}
Let $A = \mleft[\begin{smallmatrix} a & b \\ c & d \end{smallmatrix}\mright]$ be a $2 \times 2$ symplectic matrix, that is, it satisfies the equation $ad - bc = 1$. Then, the spectrum of $A$ can be written in terms of its trace, $\tau = a+d$, as
\begin{equation}\label{lambdafromtau}
\lambda_\pm = \frac{\tau \pm \sqrt{\tau^2 - 4}}2.
\end{equation}

Consequently, the behavior of the spectrum can be qualitatively described as follows
\begin{itemize}
\item If $\tau \geq 2$, both eigenvalues are positive real numbers,
\item If $-2 \leq \tau \leq 2$, both eigenvalues lie on the unit circle as a conjugate pair,
\item If $\tau \leq -2$, both eigenvalues are negative real numbers.
\end{itemize}
\end{prop}

\begin{proof}
Equation \eqref{lambdafromtau} is a trivial consequence of the fact that the characteristic polynomial of a $2 \times 2$ matrix $A$ is $p(\lambda) = \lambda^2 - (\trace A) \lambda + \det A$.

The quantitative behavior is proved in two cases.
\begin{itemize}
\item Suppose that $\abs \tau \geq 2$. Then $\lambda_\pm$ is real. Furthermore, $\lambda_{\sign \tau}$ clearly has the same sign as $\tau$, and since $\lambda_+ \lambda_- = 1$, both eigenvalues must have the same sign.
\item Suppose that $\abs \tau \leq 2$. Then, the square root term is purely imaginary and so the absolute value of $\lambda_\pm$ can be easily computed by the pythagorean theorem. Since the norm of $\lambda_\pm$ equals one and $\lambda_+ = 1/\lambda_-$, we conclude that $\lambda_+ = \conj{\lambda_-}$.
\end{itemize}
\end{proof}

\begin{corollary}\label{sp2sgn}
A matrix $A \in Sp(2)$ is in $Sp(2)^*$ iff $\trace A \neq 2$. In this case, it is in $Sp(2)^{\sign(2 - \trace A)}$.
\end{corollary}

\begin{proof}
Using the notation of proposition \ref{lambdafromtau}, since $\lambda_+ \lambda_- = 1$, the only way for either eigenvalue to be equal to 1 is for both of them to be equal to 1. Therefore, if any eigenvalue is one, $\trace A = \lambda_+ + \lambda_- = 2$. On the other hand, if $\trace A = 2$, by \eqref{lambdafromtau} both eigenvalues are one.

To determine the sign of $\det(A-I)$, we compute it directly using a known formula for the characteristic polynomial
\begin{equation}
\det(A-I) = 1^2 - (\trace A) \times 1 + \det A = 2 - \trace A.
\end{equation}
\end{proof}

\begin{remark}
Note the counter-intuitive signs. If $\trace A > 2$ then $A \in Sp(2)^-$, not $Sp(2)^+$. The counter-intuitiveness extends to the $W^\pm$ matrices, given the unfortunate fact that $\rho(W^\pm) = \mp 1$.
\end{remark}

\begin{corollary}\label{sp2pm}
Let $A \in Sp(2)$. Then, $\rho(A) = 1$ iff $A \in Sp(2) \setminus Sp(2)^+$, i.e. iff $\trace A \geq 2$. Furthermore, $\rho(A) = -1$ iff $\trace A \leq -2$.
\end{corollary}

\begin{proof}
We divide the proof in three cases.
\begin{itemize}
\item If $\trace A \geq 2$, both eigenvalues of $A$ are positive, and so using property \ref{rhodef:realev} from theorem \ref{rhodef} one concludes $\rho(A) = 1$.
\item If $\trace A \leq -2$, both eigenvalues of $A$ are negative, and so for the same reason $\rho(A) = -1$.
\item If $-2 < \trace A < 2$, both eigenvalues are in $S^1 \setminus \R$. Furthermore, they are both distinct, and so we may apply equation \eqref{rhoformula} to compute
\begin{equation}\label{sp2pm:3}
\rho(A) = \lambda_+^\sigma,
\end{equation}
where $\sigma$ is either $1$ or $-1$. Therefore, in this case $\rho(A)$ is either $\lambda_+$ or $\lambda_-$ and necessarily not a real number.
\end{itemize}
\end{proof}

The above propositions, notably corollary \ref{sp2pm}, greatly simplify the work of calculating the Maslov index in dimension 2, specifically in step \ref{maslov:step2}, which consists of extending the path $A(t)$ to one with endpoint in $W^\pm$.

\begin{corollary}\label{sp2rhoextension}
Let $A \in Sp(2)^*$, and define $A(t)$ as a path in $Sp(2)^*$ such that $A(0) = A$ and $A(1) = W^\pm$. Then, $\rho(A(t))$ has one of two possible behaviors:
\begin{itemize}
\item If $\rho(A) = 1$ then $\rho(A(t))$ is constant equal to one,
\item If $\rho(A) \neq 1$ then $\rho(A(t))$ is a path in $S^1 \setminus \{1\}$ with $\rho(A(1)) = -1$.
\end{itemize}
\end{corollary}

\begin{proof}
If $\rho(A) = 1$ then $A \in Sp(2)^-$ by corollary \ref{sp2pm}. Since the path connecting $A$ to $W^-$ remains in $Sp(2)^-$, and so $\rho(A(t))$ is constant equal to 1 by the same corollary.

If $\rho(A) \neq 1$ then $A \in Sp(2)^+$, and so $A(t)$ connects $A$ to $W^+$. Since this path does not leave $Sp(2)^+$ it never attains the value $1$, and $\rho(A(1)) = \rho(W^+) = -1$.
\end{proof}

Due to corollary \ref{sp2rhoextension}, the algorithm to calculate the Maslov index of a path of matrices can be simplified as follows.

\begin{prop}
Let $A(t)$ be a path of matrices in $Sp(2)$ with $A(T)$ in $Sp(2)^*$. Then, the Maslov index of the path $A(t)$ can be computed as follows:

\begin{enumerate}[algorithm]
\item Compute the path $\gamma(t) = \rho(A(t))$. 
\item If $\gamma(1) \neq 1$, extend it by connecting $\gamma(1)$ to $-1$, without passing through 1, otherwise leave it as-is. Call the resulting curve $\gamma_0$.
\item Compute the (clockwise) winding number of $\gamma_0^2$.
\end{enumerate}
\end{prop}

To conclude this section, we present a further way to simplify the computations. The essential observations are the following.
\begin{itemize}
\item It is trivial to compute $\rho(A)$ when $\abs{\trace A} \geq 2$, as it coincides with $\sign \trace A$,
\item $\rho(A)$ is never $\pm 1$ when $\abs{\trace A} < 2$,
\item We are only interested in the winding number of $\rho^2(A(t))$,
\item And consequently, if for some $t_0, t_1$ we have $\trace A(t_0) = - \trace A(t_1) = \pm 2$ and $-2 < \trace A(t) < 2$ for $t_0 < t < t_1$, it suffices to compute $\rho(A(t))$ for a single value of $t \in \ointerval{t_0}{t_1}$, to know whether $\rho(A(t))$ crosses the circle in the clockwise or counter-clockwise direction.
\end{itemize}

\begin{prop}\label{calcmaslov0}
Let $A(t)$, $t \in \interval 0 T$ be a path of matrices with $A(0) = I$, $A(T) = W^\pm$. If $\trace A(t) > -2$ for all time, the Maslov index of $A$ is zero. Otherwise, there exists a partition of $\interval 0 T$ of the form
\begin{equation}\label{calcmaslov:ab1}
0 = a_0 < b_0 < a_1 < b_1 < a_2 < \dots < a_{N-1} < b_{N-1} < a_N = T
\end{equation}
which satisfies the properties
\begin{enumerate}
\item\label{calcmaslov:ab2} $(-1)^n \trace A(a_n) \geq 2$, \item\label{calcmaslov:ab3} $\trace A(b_n) \in \ointerval{-2}2$, \item\label{calcmaslov:ab4} Whenever $\trace A(x) \geq 2$ and $\trace A(y) \leq -2$, there exists some $b_n$ between $x$ and $y$.\end{enumerate}

For any such partition, the negative winding number of $\rho^2(A(t))$ is equal to
\begin{equation}\label{calcmaslov0:mu}
\mu(A(t)) = \sum (-1)^n \sign(A(b_n)_{12}).
\end{equation}
\end{prop}

\begin{proof}
First we do the easy part: If $\trace A(t) > -2$ for all time, the Maslov index of $A(t)$ is null.

By corollary \ref{sp2pm}, this hypothesis implies that $\rho(A(t))$ is never $-1$. Consequently, the path $\rho(A(t))$ is contractible, and therefore so is $\rho^2(A(t))$. As such, the winding number of $\rho^2(A(t))$ is zero, and so is the Maslov index.

Now, assume that $\trace A(t) \leq -2$ for some $t$. We prove that a partition \eqref{calcmaslov:ab1} satisfying \ref{calcmaslov:ab2}, \ref{calcmaslov:ab3} and \ref{calcmaslov:ab4} exists.

Let $U \subseteq \interval 0 T$ be the preimage under $\trace \circ A$ of $\ointerval{-2}\infty$, and $V$ the preimage of $\ointerval{-\infty}2$. Then, by standard Lebesgue number lemma arguments, there exists a partition 
\begin{equation}
0 = b_{-1} < b_0 < \dots < b_{N-1} < b_N = T,
\end{equation}
such that each interval $\interval{b_n}{b_{n+1}}$ is entirely contained in $U$ or $V$. By removing unnecessary partitions, we may suppose without loss of generality that
\begin{equation}
\interval{0}{b_0} \subseteq U, \interval{b_0}{b_1} \subseteq V, \interval{b_1}{b_2} \subseteq U, \dots,
\end{equation}
and furthermore that for each $n$ there exists some $a_n \in \ointerval{b_{n-1}}{b_{n}}$ such that $a_n \in U \setminus V$ or $a_n \in V \setminus U$. Without loss of generality pick $a_0 = 0$ and $a_N = T$. (In this step, we are assuming that $0 \neq N$, which is why the case $\forall_t \trace A(t) > -2$ must be done separately.)

We now have a sequence of numbers
\begin{equation}\label{tcineq}
0 = a_0 < b_1 < a_1 < b_2 < a_2 < \dots < a_{N-1} < b_{N-1} < a_{N} = T,
\end{equation}
which we will prove satisfies properties \ref{calcmaslov:ab2}, \ref{calcmaslov:ab3} and \ref{calcmaslov:ab4}.

\begin{itemize}
\item [\ref{calcmaslov:ab2}] Let $n$ be an even number. By hypothesis, $a_n \in U \setminus V$. Consequently, $a_n \not \in V$, i.e. $\trace A(a_n) \geq -2$. The proof is analogous for $n$ odd.

\item [\ref{calcmaslov:ab3}] For $1 \leq n \leq N-1$,
\begin{equation}
b_n \in \interval{b_{n-1}}{b_n} \cap \interval{b_n}{b_{n+1}} \subseteq U \cap V,
\end{equation}
which implies that $\trace A(b_n) \in \ointerval{-2}2$.

\item [\ref{calcmaslov:ab4}] Let $\trace A(x) \geq 2$ and $\trace A(y) \leq -2$. Then, $x \in V \setminus U$ and $y \in U \setminus V$, so they cannot both be in the same interval $\interval{b_n}{b_{n+1}}$. Therefore, there must be a value of $b_n$ (with $1 \leq n \leq N-1$) between them.
\end{itemize}

We now prove that given a partition $(a_n, b_n)$ satisfying \eqref{calcmaslov:ab1} and properties \ref{calcmaslov:ab2}, \ref{calcmaslov:ab3} and \ref{calcmaslov:ab4}, formula \eqref{calcmaslov:mu} holds.

First, observe that the winding number of $\rho^2(A(t))$ can be calculated in pieces, as the path $\rho^2 \circ A$ can be decomposed into its restriction to $\interval{a_0}{a_1}$, $\interval{a_1}{a_2}$, etc., so it suffices to calculate the winding number of $\rho^2(A(t))$ as $t$ varies from $a_n$ to $a_{n+1}$.

For the sake of argument, suppose that $n$ is even. We apply property \ref{calcmaslov:ab4} and corollary \ref{sp2pm} to observe that between $a_n$ and $b_n$ the trace of $A$ is never $-2$, and consequently $\rho(A(t))$ is never $-1$. Since $S^1 \setminus \{-1\}$ is contractible, there is a homotopy between every path from $\rho(A(a_n)) = 1$ and $\rho(A(b_n)) = \exp(\pm \I \theta_n)$. Therefore, without change to the winding number of $\rho^2$, we may replace the path $\rho(A(t))$ in $\interval{a_n}{b_n}$ by a reparametrization of $\e^{\pm \I \theta}$, $\theta \in \interval 0 {\theta_n}\}$. By the same argument, the path $\rho(A(t))$ for $t \in \interval{b_n}{a_{n+1}}$ may be replaced by $\e^{\pm \I \theta}$, $\theta \in \interval{\theta_n}\pi$, and so, for $t \in \interval{a_n}{a_{n+1}}$ the path $\rho^2(A(t))$ is homotopic to $\e^{\pm 2 \I \theta}$, $\theta \in \interval 0 \pi$, with winding number $\pm 1$, where the sign is the sign of $\Im \rho(A(b_n))$.

If $n$ is odd, the previous argument works the same way, but the expression for the winding number of $\rho^2(A(t))$ has the sign swapped, i.e. it becomes $- \Im \rho(A(b_n))$. Therefore, we conclude the formula
\begin{equation}
\mu(A(t)) = -\sum (-1)^n \sign \Im \rho(A(b_n)).
\end{equation}

Note the sign swap from the definition of Maslov index. All that remains is to compute $\rho(A(b_n))$.

\begin{lemma}
Let $A$ be a $2 \times 2$ symplectic matrix with trace in $\ointerval{-2}2$. Then, $\sign \Im \rho(A) = -\sign(A_{12}) \cdot  \I$.
\end{lemma}

\begin{lemmaproof}
Let us name the entries of $A$. Suppose that
\begin{equation}
A = \begin{bmatrix}
a & b\\
c & d
\end{bmatrix}.
\end{equation}

We will construct a path $S(t)$ from $A$ to some $A'$ whose value of $\rho$ is known. Furthermore, we will ensure that $\trace S(t)$ is always in $\ointerval{-2}2$, so that $\rho(S(t))$ is never real and therefore the sign of $\Im \rho(S(t))$ never changes.

First, observe that $b \neq 0$, for otherwise $A$ would be lower-diagonal with eigenvalues equal to $a$ and $d$, both of which are real. However, we have seen (proposition \ref{lambdafromtau}) that if the trace of $A$ is in $\ointerval{-2}2$ then its eigenvalues cannot be real.

Now, consider the straight-line path from $A$ to the matrix with null trace given by
\begin{equation}
A_0 = \begin{bmatrix}
a & b\\
c-a \frac{a+d}b & -a
\end{bmatrix}.
\end{equation}

This path is obtained via continuously adding a multiple of the first line to the second line. This does not modify the determinant, hence the path so obtained is a path of symplectic matrices. Furthermore, the trace is changing linearly from $a+d \in \ointerval{-2}2$ to $0$, and so it remains in $\ointerval{-2}2$.

Since the resulting matrix $A_0$ has determinant 1, it is easy to check that it is of the form
\begin{equation}
A_0 = \begin{bmatrix}
a & b\\
-\frac{1+a^2}b & -a
\end{bmatrix}.
\end{equation}

Furthermore, any matrix of this form has null trace and determinant equal to $-1$, so any continuous deformation of $A_0$ preserves the sign of the imaginary part of $\rho$. The parameter $a$ can vary over any real value, while $b$ can only vary in $\R \setminus \{0\}$. Hence, we can find a path from $A_0$ to a matrix of the form
\begin{equation}
A' = \begin{bmatrix}
0 & \pm 1\\
\mp 1 & 0
\end{bmatrix},
\end{equation}
whose value of $\rho$ is known to be $\mp 1$ by property \ref{rhodef:xy} from theorem \ref{rhodef}. Since $\sign(b) = \pm 1$, we obtain the expected result
\begin{equation}
\sign \Im \rho(A) = - \sign(A_{12}).
\end{equation}
\end{lemmaproof}
This concludes the proof of the proposition.
\end{proof}

The previous proposition requires using a path that terminates in $W^\pm$. The following corollary shores up that weakness.

\begin{corollary}\label{calcmaslov1}
Let $A(t)$, $t \in \interval 0 T$ be a path of matrices with $A(0) = I$ and $A(T) \in Sp(2)^*$. If $\trace A(t) > -2$ for all time and $\trace A(T) > 2$, the Maslov index of $A$ is zero. Otherwise, there exists a partition of the form
\begin{equation}
0 = a_0 < b_0 < a_1 < b_1 < a_2 < \dots < a_N < b_N \leq T
\end{equation}
which satisfies the properties
\begin{enumerate}
\item $(-1)^n \trace A(a_n) \geq 2$,
\item $\trace A(b_n) \in \ointerval{-2}2$,
\item Whenever $\trace A(x) \geq 2$ and $\trace A(y) \leq -2$, there exists some $b_n$ between $x$ and $y$,
\item\label{calcmaslov:ab5} Exactly one of $\trace A(a_N)$ and $\trace A(T)$ is in $\rinterval 2 \infty$.
\end{enumerate}

For any such partition, we have the formula
\begin{equation}\label{calcmaslov:mu}
\mu(A(t)) = \sum (-1)^n \sign(A(b_n)_{12})
\end{equation}
\end{corollary}

\begin{proof}
Let $A(t)$, $t \in \interval 0 T$ be the given path of matrices, and consider an extension $\tilde A(t)$, $t \in \interval 0 {\tilde T}$, such that $\tilde A(\tilde T) = W^\pm$, with $\tilde A(t) \in Sp(2)^*$ for $t \in \interval T {\tilde T}$. We apply proposition \ref{calcmaslov0} to this new path. If $\trace A(T) > 2$ then the extension connects $A(T)$ to $W^-$, and so if $\trace A(t) > -2$ for all $t \in \interval 0 T$ we are in the conditions of the easy case of proposition \ref{calcmaslov0}. If either of these two hypotheses fail then the trace of the extension is $-2$ at some point, and so by that same proposition there is a partition
\begin{equation}
0 = a_0 < b_0 < a_1 < \dots < a_{N} < b_{N} < a_{N+1} = \tilde T.
\end{equation}

Note that, since the extension does not add any new crossings with $Sp(2)^0$, it must be the case that $a_{N} < T$.

It is tempting to throw away the last term $a_{N+1}$ to obtain the partition we seek. However, it could happen that the last term $b_N$ is greater than $T$, so we may need to perturb it, and furthermore we need to verify property \ref{calcmaslov:ab5}.

\begin{itemize}
\item Suppose that $N$ is odd, i.e. that $\trace A(a_N) \leq -2$. Since the extension does not cross $Sp(2)^0$ and $\trace A(a_{N+1}) \geq 2$, we conclude that $\trace A(T) > 2$, and furthermore that $\trace A(t) > 2$ for $t \in \interval T {\tilde T}$. As a consequence, $b_N$ is already in $\interval{a_N}T$ and requires no perturbation. It is trivial to check that properties \ref{calcmaslov:ab2}, \ref{calcmaslov:ab3} and \ref{calcmaslov:ab4} hold (they are inherited from the original partition), and the fact that $\trace A(T) > 2$ guarantees that \ref{calcmaslov:ab5} holds.

\item Suppose now that $N$ is even, i.e. that $\trace A(a_N) \geq 2$. Then, since the extension does not cross $Sp(2)^0$, we have $\trace A(T) < 2$, guaranteeing \ref{calcmaslov:ab5}. If $b_N \leq T$, it requires no perturbation and the partition obtained by throwing $a_{N+1}$ away has the desired properties.

On the other hand, suppose that $b_N > T$. If $\trace A(T) > -2$ then we change $b_N$ to equal $T$. Otherwise, if $\trace A(T) \leq -2$, we set $b_N$ to be the maximal value of $t \in \interval{a_N}T$ such that $\trace A(t) = 0$. It is straight-forward to verify that in either case the perturbed partition satisfies the necessary properties \ref{calcmaslov:ab2} to \ref{calcmaslov:ab5}.
\end{itemize}

Note that from such a partition $a_0 < b_1 < \dots < b_N$ we can recover the partition $a_0 < b_1 < \dots < a_{N+1}$ required to apply proposition \ref{calcmaslov0} to $\tilde A(t)$, by defining $a_{N+1} = \tilde T$. Property \ref{calcmaslov:ab5} is necessary to ensure that $\trace A(a_{N+1})$ has the right sign. Applying formula \eqref{calcmaslov0:mu} we may compute the negative winding number of $\rho^2(\tilde A(t))$, and consequently the Maslov index of $A(t)$, yielding the formula \ref{calcmaslov:mu}.
\end{proof}

\subsection{The Autonomous Case}

Let $H$ be an autonomous Morse function on a symplectic manifold $M$ whose time-$T$ flow $\phi_T$ is a nondegenerate Hamiltonian diffeomorphism. In this section, we will compute the Maslov index of the critical points of $H$ and relate them to their Morse index. At the end, we will repeat the procedure for the two-dimensional case, as an example of application of the methods from section \ref{sec:maslovsp2}.

Let $x$ be a critical point of $H$, i.e. $(\dl H)_x = 0$. Then, the periodic time-$T$ orbit starting at $x$ is precisely the constant orbit. We will show that this orbit is nondegenerate, so that we may compute its Maslov index.
[...]

[Some theorem about in critical points $x$ of $H$, the path $A$ can be chosen to be $\exp(t J_0 S)$, where $S$ is the Hessian of $H$ at $x$ written in any basis.]

As a consequence, if $Z_1, \dots, Z_{2n}$ is a basis of $T_x M$, we may consider $A(t)$ as the representation of $(\dl \phi_t)_x$ in this basis (see steps \ref{maslovalg:step1} to \ref{maslovalg:step5}).

[Temporary? Kick the can up the road and just refer to Audin and Damian]

\begin{theorem}
If $S$ is an invertible symmetric matrix with operator norm $\norm S < \frac{2\pi}T$ and if $A(t) = \exp(t J_0 S)$, $t \in \interval 0 T$, then $A(T)$ is nondegenerate and
\begin{equation}
\mu(A(t)) = \Ind(S) - n,
\end{equation}
where $\Ind$ denotes the Morse index.
\end{theorem}

\begin{proof}
See \cite{audin}, proposition 7.2.1.
\end{proof}

\subsection{The Two-Dimensional Case}

We will now use the tools developed in section \ref{sec:maslovsp2} to recompute the Maslov index of the critical points of $H$.

[Nondegeneracy?]

We have already shown that the Maslov index of such a critical point is given by the Conley-Zehnder index of the path
\begin{equation}
A(t) = \exp(t J_0 S), t \in \interval 0 T,
\end{equation}
with $S = \hessian H(x)$, i.e. $S_{ij} = \hessian H(x)(Z_i, Z_j) = Z_i \cdot (Z_j \cdot H)$. It is therefore to such a path that we will focus our attention.

It is fortunate that the theory greatly simplifies in the two-dimensional case, allowing us to do the computations by hand.

\begin{prop}\label{prop:sidentities}
Let $S$ be a $2 \times 2$ symmetric matrix, and let $\Delta = \det S$. Then, the following identities hold.
\begin{gather}
J_0 S J_0 S = - \Delta I, \label{eq:j0sj0s}\\
\exp(t J_0 S) = \begin{cases}
\cos(\sqrt{\Delta} t) I + \frac1{\sqrt{\Delta}} \sin(\sqrt{\Delta} t) J_0 S, & \Delta > 0,\\
I + t J_0 S, & \Delta = 0,\\
\cosh(\sqrt{-\Delta} t) I + \frac1{\sqrt{-\Delta}} \sinh(\sqrt{-\Delta} t) J_0 S, & \Delta < 0.
\end{cases} \label{eq:exptj0s}
\end{gather}
\end{prop}

\begin{proof}
The proof of \eqref{eq:j0sj0s} is simple computation. Using \eqref{eq:j0sj0s}, the identity \eqref{eq:exptj0s} can be obtained simply by expanding the Taylor series of the exponential and grouping terms as follows
\begin{equation}
\begin{aligned}
\exp(t J_0 S) &= 
I \color{black!60} + t J_0 S \color{black} + \frac12 t^2 (J_0 S)^2 \color{black!60} +\frac1{3!} t^3 (J_0 S)^3 \color{black} \cdots \\
&= 
I \color{black!60} + t J_0 S \color{black} - \frac12 t^2 \Delta I \color{black!60} -\frac1{3!} t^3 \Delta J_0 S \color{black} \cdots
\end{aligned}
\end{equation}

Equation \eqref{eq:exptj0s} should now be evident: the black terms correspond to cosines or hyperbolic cosines and the gray terms correspond to sines or hyperbolic sines.
\end{proof}

Proposition \ref{prop:sidentities} gives us full knowledge about $\exp(t J_0 S)$ for $S = \hessian H(x)$, and so we may use it to compute the Maslov index of this path.

\begin{prop}\label{maslovmorse}
Let $H$ be a Morse function and $x$ a critical point of $H$. Let $S = \hessian H(x)$. Then, the Maslov index $\mu(x)$ of the (constant) periodic orbit $x$ depends on the following way on the Morse index of $S$ and on the period $T > 0$:
\begin{itemize}
\item If $\Ind S = 1$ then $\mu(x) = 0$,
\item If $\Ind S = 1 \pm 1$ then the periodic orbit is nondegenerate if and only if $T \not \in \frac{2 \pi}{\sqrt\Delta} \Z$. In this case, the Maslov index is $\pm(2N+1)$, where
\begin{equation}
T \in \ointerval{\frac{2\pi}{\sqrt\Delta}N}{\frac{2\pi}{\sqrt\Delta}(N+1)}.
\end{equation}

Diagramatically, this can be represented as
\begin{figure}[H]
\centering
\begin{tikzpicture}
\draw[->,thick] (0,0)--(10.6,0) node[below left] {$T$};
\draw[] (0,-0.200)--(0,0.200) node[above] {0};
\foreach \i in {2,4,...,10}
\draw[] (\i,-0.200)--(\i,0.200) node[above] {$\displaystyle \frac{\i\pi}{\sqrt\Delta}$};


\foreach \i in {1,3,...,9}
\draw[] (\i,0) node[below] {$\mu = \pm \i$};
\end{tikzpicture}
\caption{Representation of the Conley-Zehnder index of the path $\exp(t J_0 S)$, $t \in \interval 0 T$ as a function of $T > 0$.}
\end{figure}
\end{itemize}
\end{prop}

\begin{proof}
We begin by solving the case $\Ind S = 1$. By inspection of equation \eqref{eq:exptj0s}, if $\Ind S = 1$ then both eigenvalues have distinct signs and $\det S = \Delta < 0$. Therefore, by \eqref{eq:exptj0s},
\begin{equation}
\trace \exp(t J_0 S) = 2 \cosh(\sqrt{-\Delta} t) \geq 2,
\end{equation}
where we used the fact that $\trace(J_0 S) = 0$. Application of corollary \ref{calcmaslov1} yields that, in this case, $\mu(x) = 0$.

We now focus on the case $\Ind S = 1 \pm 1$. In this case, by a similar calculation we obtain $\Delta > 0$ and furthermore
\begin{equation}
\trace \exp(t J_0 S) = 2 \cos(\sqrt\Delta t).
\end{equation}

Therefore, by corollary \ref{sp2sgn} the path $\exp(t J_0 S)$ is degenerate iff $\cos(\sqrt \Delta T) = 1$, which happens iff $\sqrt\Delta T \in 2 \pi \Z$. This completes the proof of nondegeneracy.

\smallskip

To compute the Maslov index, we apply corollary \ref{calcmaslov1}. We have already seen that $\trace \exp(t J_0 S) = 2 \cos(\sqrt\Delta t)$, and so we may build the partition required by corollary \ref{calcmaslov1} explicitly. The interval $\interval 0 T$ can be divided into intervals of size $\frac\pi{\sqrt\Delta}$, on the extrema of which the trace is $\pm 2$. Therefore, we set $a_n = n \frac\pi{\sqrt\Delta}$. We may pick any $b_n$ between $a_n$ and $a_{n+1}$; for example their midpoints, on which the trace is zero. This ensures that properties \ref{calcmaslov:ab2} and \ref{calcmaslov:ab3} hold.

Note that (for all $T$ such that our path is non-degenerate) we have $\trace \exp(t J_0 S) \in \rinterval{-2}2$, so that in order to verify property \ref{calcmaslov:ab5} the last element of our sequence $a_n$ needs to satisfy $\trace \exp(a J_0 S) = 2$, and so we terminate our sequence at $a_{2N}$, even if $a_{2N+1}$ happens to be less than $T$. Finally, we pick $b_{2N}$ arbitrarily, as long as it is between $a_{2N}$ and $\min(T, a_{2N+1})$.

In summary, we consider the partition
\begin{equation}\label{autmaslov:thepartition}
\begin{array}{c|c|c|c|c|c|c|c}
a_0 & b_0 & a_1 & b_1
& \cdots &
b_{2N-1} & a_{2N} & b_{2N} \\
\hline
0 & \frac12 \frac{\pi}{\sqrt\Delta} & \frac{\pi}{\sqrt\Delta} & \frac32 \frac{\pi}{\sqrt\Delta}
& \cdots &
\frac{2(2N-1)+1}2 \frac{\pi}{\sqrt\Delta} & 2N \frac{\pi}{\sqrt\Delta} & b_{2N}
\end{array}
\end{equation}

We have already verified most of the conditions of corollary \ref{calcmaslov1}, the only one remaining being property \ref{calcmaslov:ab4}, i.e. that whenever $\trace A(x) \geq 2$ and $\trace A(y) \leq -2$ there exists some $b_n$ between $x$ and $y$. This property is verified directly by checking that in this event $x$ is $a_n$ for some even $n$ and $y$ is $a_n$ for some odd $n$, and by construction there is always a $b_n$ between such numbers.

Corollary \ref{calcmaslov1} now tells us that the Maslov index of $A(t) = \exp(t J_0 S)$ equals
\begin{equation}
\mu(A(t)) = \sum_{n = 0}^{2N} (-1)^n \sign(A(b_n)_{12}).
\end{equation}

We compute $A(b_n)$, for $n < 2N$ explicitly as
\begin{equation}
A(b_n) = \frac1{\sqrt\Delta} \sin\left( \sqrt\Delta \frac{2n + 1}2 \frac\pi{\sqrt\Delta} \right) J_0 S = (-1)^n J_0 S,
\end{equation}
and therefore $\sign(A(b_n)_{12}) = (-1)^n \sign((J_0 S)_12)$. The case $n = 2N$ requires a little extra consideration, but it is easy to verify that the conclusion is the same, and so we have the identity
\begin{equation}
\mu(A(t)) = (2N+1) \sign((J_0 S)_{12}),
\end{equation}
and so it remains to relate the sign of $(J_0 S)_{12}$ to the index of $S$.

First, we make the trivial observation that $(J_0 S)_{12} = - S_{22}$, so it suffices to relate the sign of $S_{22}$ to the index of $S$.

Furthermore, observe that $S_{22}$ and $S_{11}$ have the same sign, as
\begin{equation}
S_{11} S_{22} = \Delta + S_{12}^2 > 0.
\end{equation}

Finally, this sign coincides with the sign of both eigenvalues of $S$ (which have the same sign because $\lambda_1 \lambda_2 = \Delta > 0$), and both of these signs equal the sign of the trace of $S$. In conclusion,
\begin{itemize}
\item The sign of $S_{22}$ equals the sign of the trace of $S$, which equals the sign of both eigenvalues of $S$.
\item Consequently, if $\Ind S = 0$ then these eigenvalues are positive, and therefore so is $S_{22}$ and therefore $(J_0 S)_{12} < 0$ and $\mu(A(t)) = -(2N+1)$.
\item On the other hand, if $\Ind S = 2$ then these eigenvalues are both negative, and by the same argument $\mu(A(t)) = 2N+1$.
\end{itemize}

The proof is therefore complete.
\end{proof}

We note that the usual result on $C^2$-small Morse functions can be recovered.

\begin{corollary}
Let $H$ be an autonomous Morse function on a 2-dimensional compact symplectic manifold $M$. Then, for small enough $\varepsilon > 0$, the function $\varepsilon H$ is a nondegenerate Hamiltonian, i.e. the time-one flow of $\varepsilon H$ is a nondegenerate Hamiltonian diffeomorphism. More precisely, the only periodic (time-one) orbits of $\varepsilon H$ are constants equal to critical points of $H$. In this case, the action functional of these constant orbits coincides with their value of $\varepsilon H$, and their Maslov index and Morse index are related by the formula
\begin{equation}
\mu(x) = \Ind(x) - 1.
\end{equation}
\end{corollary}

\begin{proof}
First we prove that, for small enough $\varepsilon$, the only time-one orbits of $\varepsilon H$ are the constants. Equivalently, it suffices to show that the periods of nonconstant orbits of $H$ are bounded from below, as the time-one orbits of $\varepsilon H$ coincide with the time-$\varepsilon$ orbits of $H$.

[Write this sometime, or refer to \cite{audin}, proposition 6.1.6]

Assume now that $\varepsilon$ is small enough that the only time-one orbits of $\varepsilon H$ are constants. Recall the definition of the action functional on a periodic orbit $x$,
\begin{equation}
\AA_{\varepsilon H}(x) = - \int_D \omega + \int_0^1 \varepsilon H(x(t)) \dl t,
\end{equation}
where $D$ is any map from the disk to $M$ whose border is $x$.

Since $x$ is constant, we might as well take $D$ to be the constant map. This makes the pullmack of $\omega$ to the disk be null, and so the first term from the action functional vanishes. As for the second term, since $H$ is autonomous and $x$ is constant, we obtain simply
\begin{equation}
\AA_{\varepsilon H}(x) = \varepsilon H(x).
\end{equation}

Finally we show that, possibly by decreasing $\varepsilon$, the Maslov index of these constant orbits coincides up to a shift by one with their Morse index.

Application of propositions \ref{to do} and \ref{maslovmorse} will show that in order for the relation $\mu(x) = \Ind(x) - 1$ to be verified it suffices that for every critical point $x$ of $H$ there exists a symplectic basis for $T_x M$ such that the determinant of the Hessian of $\varepsilon H$ in that basis\footnote{It is a trivial observation, yet unnecessary for our purposes, that the determinant of the Hessian of a smooth function is independent of the symplectic basis chosen.} is less than $4 \pi^2$. A simple computation will show that, for this effect, it suffices that
\begin{equation}
\varepsilon < \min \left( \frac{2\pi}{\sqrt{\det \hessian H(x)}}\right),
\end{equation}
where the minimum ranges over the critical points of $H$.
\end{proof}

A somewhat surprising corollary of proposition \ref{maslovmorse} is that the Maslov indices of critical points of autonomous Hamiltonians may only take odd values or zero. As a consequence of this and the fact that Floer homology must coincide with Morse homology, if any critical point of $H$ has Maslov index different from $-1$, $0$, and $1$, then the time-one flow of $H$ \emph{must} be degenerate. This allows us to prove the existence of orbits with almost all possible non-small periods.

\begin{corollary}
Let $H$ be a Morse function on the two-dimensional compact symplectic manifold $M$. Let $x_1, \dots, x_N$ be the critical points of $H$ \emph{with Morse index different from $0$}, i.e. the local maxima and minima.

For each $i = 1, \dots, N$, define $\Delta_i = \det \hessian H(x_i)$, where the determinant is taken in an arbitrary symplectic basis of $T_{x_i} M$, and set $q_i = \frac{2\pi}{\sqrt{\Delta_i}}$ and $q_{\min} = \min q_i$. Then, for all $p \geq q_{\min}$, the time-$p$ flow of $H$ is a degenerate Hamiltonian diffeomorphism; equivalently, $p H$ is a degenerate Hamiltonian.

Furthermore, if we let
\begin{equation}
A = \ointerval{q_{\min}}{\infty} \setminus \bigcup q_i \Z,
\end{equation}
then for all $p \in A$ there exists a time-$p$ nonconstant periodic orbit of $H$, though $p$ may in general not be a primitive period.

[Should be able to prove that there exist orbits with periods greater than some $T$ constructed from the $\Delta_i$.]
\end{corollary}

\begin{proof}
Consider the Hamiltonian $p H$, $p > 0$. It has the same critical points as $H$, with the same indices, and the determinants of the Hessians of the $x_i$ can be easily computed as
\begin{equation}
\det \hessian(pH)(x_i) = p^2 \Delta_i.
\end{equation}

We conclude from proposition \ref{maslovmorse} that if $p \geq \frac{2\pi}{\sqrt{\Delta_i}} = q_i$ then either the critical point $x_i$ is degenerate (in the Floer sense) or or it has Maslov index in $\Z \setminus \{-1,0,1\}$. Therefore, if $p \geq q_{\min}$ either some critical point is degenerate (in which case the Hamiltonian $pH$ is degenerate) or some critical point has Maslov index in $\Z \setminus \{-1,0,1\}$. We will now show that the latter also implies that $pH$ is degenerate.

Assume for the sake of argument that $pH$ is a nondegenerate autonomous Hamiltonian which has a critical point $x$ with Maslov index $\mu(x) = m \not \in \{-1,0,1\}$. Then, the Floer complex is nontrivial at degree $m$, i.e. $\CF_m(pH) \neq \{0\}$. Therefore, since the Morse homology agrees with Floer homology, the boundary maps $\partial_m$ and $\partial_{k+1}$ must satisfy $\ker(\partial_m) = \image(\partial_{k+1})$. However, proposition \ref{maslovmorse} implies that no critical point can have Maslov index equal to $m \pm 1$, and therefore that $\CF_{m\pm1}(pH) = \{0\}$. As a consequence, $\ker(\partial_m) = \CF_m(pH)$ and $\image(\partial_{k+1}) = \{0\}$, hence $\HF_m(pH) = \CF_m(pH) \neq \{0\}$. This is a contradiction, because the Morse homology at index $m+1$ is necessarily null, as Morse indices may only take the values $0$, $1$, and $2$. This shows that $pH$ is necessarily degenerate for $p \geq q_{\min}$.

\smallskip

We now turn to showing that if $p \in A$ then there necessarily exists a $p$-periodic orbit of $H$. Indeed, as we have seen, if $p \in A$ then the Hamiltonian $pH$ is degenerate. However, if $p$ is not a multiple of $q_i$ for any $i$, then by proposition \ref{maslovmorse} every critical point of $H$ is a nondegenerate orbit. Therefore, there must be a degenerate time-one orbit of $pH$ which is not a constant, or, in other words, a non-constant time-$p$ orbit of $H$.
\end{proof}

\subsection{Relation to a construction by Hofer and Kriener}

There is another construction of the Conley-Zehnder index, also specific to the 2-dimensional case, due to Hofer and Kriener. The original work can be found in section 3 of \cite{hoferkriener}, and a summary can be found in the appendix of \cite{hwz}. For convenience, we reproduce the construction.

Let $A(t)$, $t \in \interval 0 T$, be the path of $2 \times 2$ symplectic matrices whose Conley-Zehnder we intend to compute. Given a nonzero complex number $z \in \C \cong \R^2$. we can consider the path of nonzero complex numbers given by $z(t) = A(t) z$. We may write this path in polar coordinates as
\begin{equation}
z(t) = \abs{z(t)} \exp(\I \theta(t)),
\end{equation}
where $\theta(t)$ is chosen to be continuous. Then, we define the winding of $z$ by
\begin{equation}
\Delta(z) = \theta(T) - \theta(0).
\end{equation}

Finally, we define the winding interval of $A(t)$ as
\begin{equation}
I(A) = \{\, \Delta(z) \mid z \in \C \setminus \{0\} \,\}.
\end{equation}

It will be shown that this set is a closed interval of length strictly less than $\pi$, and therefore it contains at most one multiple of $2\pi$. Then, we may define the CZ index \textit{a la} Hofer and Kriener as
\begin{equation}
\muhk(A) = \begin{cases}
2 k, & 2 k \pi \in I(A),\\
2k+1, & I(A) \subseteq \ointerval {2k\pi}{2(k+1)\pi}.
\end{cases} 
\end{equation}

We begin by verifying the statements made above, in order to ensure that $\muhk$ is well-defined.

Recall that the function $f(t) = \exp(\I t)$ is a covering map $\R \xrightarrow{f} S^1$. Therefore, the following proposition guarantees that for each $z$ a continuous argument for $z(t)$ can be found. [Is this necessary?]

\begin{prop}[Path Lifting Property]
Let $p \colon E \to B$ be a covering map and $\gamma \colon \interval01 \to B$ a curve. Then, for every $e \in p^{-1}(\gamma(0))$ there exists a unique continuous lift $\tilde \gamma \colon \interval01 \to E$ satisfying $\gamma = p \circ \tilde \gamma$.
\end{prop}

\begin{proof}
See proposition 11.10 in \cite{leetopological}.
\end{proof}

\begin{corollary}
Any continuous path $z(t)$ of nonzero complex numbers has a continuous argument. For any such choice, the winding of $z(t)$ is the same.
\end{corollary}

\begin{proof}
To obtain the existence of a continuous argument, apply the path lifting property to the function $t \mapsto \frac{z(t)}{\abs{z(t)}}$, using the covering map $\exp(\I\,\cdot\,) \colon \R \to S^1$ and any value of $\theta_0 \in \R$ satisfying $\exp(\I \theta_0) = z(0)$.

Now, we show that the value of $\Delta(z) = \theta(T) - \theta(0)$ does not depend on the chosen argument. Therefore, let $\theta'$ be another continuous argument of $z(t)$. Then, $\theta'(0)$ must be equal to $\theta(0) + 2 k \pi$ for some integer $k$, as $\exp(\I\theta'(0)) = z(0) = \exp(\I \theta(0))$. Consequently, we can show that $\theta'(t) = \theta(t) + 2 k \pi$, using the uniqueness of the lift relative to the starting point. As a consequence,
\begin{equation}
\theta'(T) - \theta'(0) = \theta(T) + 2 k \pi -  \theta(0) - 2 k \pi = \theta(T) - \theta(0),
\end{equation}
so that this value is independent of the chosen lift.
\end{proof}

Now that we have shown that $\Delta(z)$ is well-defined, we show that it is a continuous function of $z$. To do so, we apply a stronger lifting property.

\begin{prop}[Lifting Criterion]
Suppose $p \colon E \to B$ is a covering space. Let $X$ be a connected and locally path connected space, and let $f \colon X \to B$ be a continuous map. Given any points $x_0 \in X$ and $e_0 \in E$ such that $p(e_0) = f(x_0)$, $f$ has a lift $\tilde f \colon X \to E$ satisfying $\tilde f(x_0) = e_0$ if and only if $f_* \pi(X,x_0) \subseteq p_* \pi(E,e_0)$, where $\pi$ denotes the fundamental group.

If this lift exists, it is unique.
\end{prop}

\begin{proof}
For the proof of the first part, see theorem 11.15 in \cite{leetopological}.

To prove uniqueness, we apply the path lifting property. It is a known fact from topology that if $X$ is connected and locally path connected then it is connected. As such, for any $x \in X$ there exists a path $\gamma$ from $x_0$ to $x$.

If $\tilde f$ is a lift of $f$, then it is easy to check that $\tilde f \circ \gamma$ is a lift of the path $f \tilde \gamma$ which starts at $e_0$. The path lifting property guarantees that this lift is unique, and so $\tilde f(\gamma(t))$ is uniquely defined for every value of $t$. In particular, this determines $\tilde f(x)$, and since $x$ is arbitrary the lift $\tilde f$ is completely determined, i.e. such a lift is unique.
\end{proof}

\begin{prop}
The winding number function $\Delta \colon \C \setminus \{0\} \to \R$ is continuous.
\end{prop}

\begin{proof}
To help in the proof, we define $\Delta$ in an alternate way, using the lifting properties above, in such a way that it will be clear that $\Delta$ is continuous. In what follows, set $\C^* = \C \setminus \{0\}$.

First, we consider the continuous function
\begin{equation}
\begin{aligned}
f_0 \colon \C^* \times \interval0T &\to S^1\\
(z,t) &\mapsto \frac{A(t) z}{\abs{A(t) z}}
\end{aligned}
\end{equation}
and then set $f(z,t) = f_0(z,t) / f_0(z,0)$. Note that $f(z,0) = 1$ for all $z \in \C^*$.

We intend to lift $f$ to a map $\tilde f \colon \C^* \times \interval0T \to \R$, using the lifting criterion above. For concreteness, we require that $\tilde f(1,0) = 0$. We apply the lifting criterion. Because $\C^* \times \interval0T$ satisfies all the connectivity assumptions, we need only verify the condition
\begin{equation}
f_* \pi(\C^* \times \interval0T, (1,0)) \subseteq p_* \pi(\R,0).
\end{equation}

Since $\R$ is simply connected, the right-hand side is the null group and so it is necessary to prove that
\begin{equation}\label{eq:fstarpi0}
f_* \pi(\C^* \times \interval0T, (1,0)) = \{0\}. 
\end{equation}

To do so, note that $\C^* \times \interval01$ can be retracted by deformation to $\C^* \times \{0\}$, and so every element of the fundamental group of $\C^* \times \interval0T$ has a representative in $\C^* \times \{0\}$. This set is mapped by $f$ to $\{1\} \subseteq S^1$, which proves that $f_*(\alpha)$ is homotopic to the constant path, for any $\alpha \in \pi(\C^* \times \interval0T,(1,0))$, and so we have proven \eqref{eq:fstarpi0}. Hence the lifting criterion is applicable and $f$ can be lifted to a map
\begin{equation}
\tilde f \colon \C^* \times \interval0T \to \R.
\end{equation}

We remark that, by continuity, $\tilde f(z,0) = 0$ for all $z$. Indeed, since $f(z,0) = 1$ for all $z$, $\tilde f(z,0) \in 2 \pi \Z$. Since this is a discrete space and $\C^*$ is connected, all values of $\tilde f(z,0)$ must be the same, and by hypothesis $\tilde f(1,0) = 0$.

Finally, we prove that
\begin{equation}\label{eq:deltatildef}
\Delta(z) = \tilde f(z,T).
\end{equation}

To prove \eqref{eq:deltatildef}, it suffices to show that, if $\theta_0$ is an argument of $z$, then $\theta_0 + \tilde f(z,t)$ is an argument of $z(t)$. This is a trivial computation:
\begin{equation}
\begin{aligned}
\exp(\I[\theta_0 + \tilde f(z,t)]) &= \exp(\I \theta_0) \exp(\I \tilde f(z,t))\\
&= \frac z {\abs z} p(\tilde f(z,t))\\
&= f_0(z,0) f(z,t)\\
&= f_0(z,t)\\
&= \frac{z(t)}{\abs{z(t)}}.
\end{aligned}
\end{equation}

This completes the proof of \eqref{eq:deltatildef}, and therefore that $\Delta$ is a continuous function of $z$.
\end{proof}