\documentclass{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{mathtools}

\usepackage[thmmarks, amsmath]{ntheorem}

\usepackage{fullpage}

\usepackage[inline]{enumitem}


\usepackage[cal=euler]{mathalpha}
%\usepackage{ dsfont }

\usepackage{tikz}
\usepackage{float}

\setlist[enumerate,1]{label=(\alph*)}

\title{Possible Topic Proposal\\The Kueker Conjecture}
\author{Duarte Maia}
%\date{}

\theorembodyfont{\upshape}
\theoremseparator{.}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{prop}[theorem]{Proposition}
\renewtheorem*{prop*}{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\newtheorem{conjecture}[theorem]{Conjecture}

\theoremsymbol{\ensuremath{\square}}
\newtheorem{prelimdef}[theorem]{Preliminary Definition}
\newtheorem{definition}[theorem]{Definition}

\theoremseparator{:}
\newtheorem{principle}{Principle}

\theoremstyle{nonumberplain}
\theoremsymbol{}
\newtheorem{convention}{Convention}

\theoremheaderfont{\itshape}
\theorembodyfont{\upshape}
\theoremseparator{:}
\theoremsymbol{\ensuremath{\blacksquare}}
\newtheorem{proof}{Proof}
\newtheorem{explanation}{Explanation}
\theoremsymbol{\ensuremath{\text{\textit{(End proof of lemma)}}}}
\newtheorem{lemmaproof}{Proof of Lemma}

\theoremsymbol{\ensuremath{\square}}
\newtheorem{sketch}{Proof Sketch}

\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}

\newcommand{\Lang}{\mathcal{L}}
\newcommand{\calN}{\mathcal{N}}
\newcommand{\Stone}{\mathrm{S}}
\newcommand{\PStone}{\mathrm{PS}}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\PTr}{PTr}
\DeclareMathOperator{\Th}{Th}
\DeclareMathOperator{\Ext}{Ext}

\DeclarePairedDelimiter{\braket}{\langle}{\rangle}
\DeclarePairedDelimiter{\abs}{\lvert}{\rvert}



\begin{document}
\maketitle

%\tableofcontents

\section{Introduction}

This document is a brief, short-form summary of a possible Topic proposal.

\section{The Proposal Proposal}

Consider the following topics, roughly in order:

\begin{itemize}
\item Brief explanation of: Language, Theory, Model. Completeness and compactness. Some simple examples to be used later.
\item Definition of partial type. Partial types can always be realized.
\item Partial type using parameters from model. Cute application: There exist nonstandard models of PA.
\item Partial types may sometimes be omitted. Definition of $\omega$-saturation, and a brief mention of $\kappa$-saturation for higher $\kappa$. Examples and non-examples.
\item Manifestations of saturation: Morley's Thesis, possibly Isabella's work (EC groups are just $\omega$-saturated models of Grp?), sketch characterization of $\aleph_0$-categorical theories, no theory has exactly two countable models.
\item (The previous point probably requires mentioning atomicness.)
\item Remark that there seems to be a relation between saturation and counting models.
\item Enunciate Kueker's conjecture.
\item Sketch proof of the converse.
\item Talk about state of the art.
\item As part of the previous point, introduce stability. Possibly talk a little more about the concept?
\end{itemize}

%\bibliographystyle{plain}
%\bibliography{bibliography}

\end{document}