\documentclass[addpoints]{exam}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{tikz}

\DeclarePairedDelimiter{\Norm}\lVert\rVert
\newcommand{\R}{\mathbb{R}}
\newcommand{\tr}{\intercal}

%EDIT THESE WHEN NEW CLASS
\newcommand{\themath}{19620}
\newcommand{\thesec}{46}

\pagestyle{headandfoot}
\runningheader{MATH \themath\ Section \thesec}{Midterm 1 Sample}{Duarte Maia}
\runningheadrule
%\footer{}{\ifincomplete{Exercise continues on next page.}{}}{}


\begin{document}

\begin{center}
\Large Quiz 3

MATH \themath - Section \thesec

\smallskip

\large Section Instructor: Duarte Maia

\smallskip

February 19, 2026
\end{center}

\noindent
\makebox[\textwidth]{
Name and UCID: \,\hrulefill
}

\begin{center}
\fbox{
\parbox{6in}{
\begin{itemize}
\item You have 45 minutes to answer the following question in the space given below.

\item You may ask for additional paper if necessary.

\item Your solution should use methods similar to those you used on the project.

\item You do not need to explain or justify your solution. However, you do need to \textbf{show your work}. Someone who knows how to solve this exercise should be able to understand how you reached your solution. An incomplete solution with full work shown is better than a correct solution with no work shown.
\end{itemize}
}
}
\end{center}

\hrule

\qformat{\textbf{Exercise:} \quad}

\begin{questions}
\question Paloma has been baking a lot of cakes over the weekend. So many, that she lost count! Your goal is to help Paloma figure out how many cakes she baked.

You are given the following information:
\begin{itemize}
\item Paloma only baked three types of cakes: Devil's Cake, Royal Cake, and German cake.
\item Devil's cake uses 2 cups of flour, 2 cups of sugar, 0.75 cups of cocoa, and 2 eggs.
\item Royal cake uses 0.75 cups of flour, 4 cups of sugar, 1 cup of cocoa, and 4 eggs,
\item German cake uses 2.5 cups of flour, 2 cups of sugar, 2 cups of cocoa, and 4 eggs,
\item Paloma used up \underline{about} one and a half bags of flour, each about 18 cups, ($1.5\times 18 = 27$)
\item Paloma used up \underline{a bit under} six bags of sugar, each about 9 cups, ($6\times 9 = 54$)
\item Paloma used up \underline{almost} 10 cans of cocoa powder, each about 2 cups, ($10\times 2 = 20$)
\item Paloma used up five dozen eggs. But she had butter fingers. Some of them may have fallen on the floor. She does not remember how many. But at least one. ($5\times 12 = 60$)
\end{itemize}
\begin{parts}
\part[25] Write the least-squares system that you would use in this scenario. You may write it in terms of matrix multiplication. You do not need to actually multiply any matrices.
\part[75] Let's suppose that the solution to the system you just wrote is given by:
\[\text{devil's cake} = 6.34, \quad \text{royal cake} = 8.75, \quad \text{german cake} = 3.12.\]

With this information, find out how many cakes of each kind Paloma baked over the weekend. There is a single true and correct answer. Make a real effort to find it, and cross-check your answer with Paloma's information.
\end{parts}
\end{questions}
\end{document}