% !TeX root = Thesis.tex
\chapter{Barcodes from Filtered Floer Homology}\label{chap:bffh}

\section{Introduction to Floer Homology}

Let $(M, \omega)$ be a compact symplectic manifold. We assume that $M$ is aspherical, which means that any continuous map $S^2 \to M$ is nulhomotopic. Let $\LL M$ be the space of smooth contractible loops on $M$, i.e. smooth maps $S^1 \to M$ which can be extended to the disk. By the asphericity condition any such extensions are homotopic.

Let $H$ be a smooth Hamiltonian on $M$. Through it, define the so-called action functional on $\LL M$. For $x \in \LL M$, set
\begin{equation}\label{eq:defaction}
\AA(x) = - \int_D \omega + \int_{S^1} H \circ x \dl3 t,
\end{equation}
where $D$ is any extension of $x$ to the disk. The value of $\AA$ does not depend on the extension by the Stokes theorem and the fact that $\omega$ is closed.

It is possible to visualize $\LL M$ as a manifold and the action functional as a Morse function on this `manifold', and hence do Morse theory on it. With this in mind, it is useful to first intuit what the derivative of $\AA$ should be.

In this metaphor, a `tangent vector to $x$' is a vector field $\xi$ on $x$, that is, a smooth collection of vectors $\xi(t) \in T_{x(t)} M$. With that, it is possible to compute
\begin{equation}
 (\dl \AA)_x(\xi) = - \int_{S^1} \xi \into \omega + \int_{S^1} (\dl H)_{x(t)} (\xi(t)) \dl3t. 
\end{equation}

Therefore, $x\in \LL M$ is a `critical point of $\AA$' if and only if $(\dl \AA)_x(\xi) = 0$ for all $\xi$, and it is straight-forward to see that this happens iff $x$ satisfies the equation
\begin{equation}
(\dl H)_{x(t)} = - \dot x(t) \into \omega.
\end{equation}

Consequently, the critical points of $\AA$ correspond to the periodic orbits of $H$. It is possible to define a notion of non-degenerate critical points, and therefore of `Morse action functional'. Since the action functional is obtained from the Hamiltonian, this non-degeneracy condition is attributed to the Hamiltonian. A Hamiltonian which induces a `Morse action functional' is said to be non-degenerate. The definition of what makes a Hamiltonian degenerate or nondegenerate is not relevant for the time being; it is discussed below when we talk about the Maslov index in chapter \ref{chap:maslov}.

If $H$ is a nondegenerate Hamiltonian, the only two ingredients left to be able to do Morse theory on the space of loops are a grading of the orbits (this is called the Maslov index; see chapter \ref{chap:maslov}) and an equivalent to a Riemannian structure, so that we may take (negative) gradients of functionals. A standard way to establish a Riemannian metric on a symplectic manifold is to pick an almost complex structure $J$ which is compatible with $\omega$, and set $\braket{v,w} := \omega(v, Jw)$. With this metric, one can also establish a `Riemannian metric' on the space of loops:
\begin{equation}
\braket{\xi, \zeta} = \int_{S^1} \braket{\xi(t), \zeta(t)} \dl3t,
\end{equation}
with which the gradient of $A$ can be written as
\begin{equation}
\grad \AA(x) = J(\dot x - X^H),
\end{equation}
where $X^H$ satisfies $X^H \into \omega = - \dl H$. As a consequence, we may write the `ODE' necessary for a `path of orbits'
\begin{equation}
u(t,s) \colon S^1 \times \R \to M
\end{equation}
to be following the flow of the negative gradient of $\AA$. This results in the PDE
\begin{equation}
\diffp u s + J \diffp u t = J X^H.
\end{equation}
A function $u$ which satisfies this PDE is said to be a pseudo-holomorphic curve.

We may now define the so-called Floer chain complex
\begin{equation}
\CF_k(M, H) := \braket{\text{periodic orbits of $H$ of degree $k$}},
\end{equation}
where the angled brackets denote the free vector space whose basis is the set inside the brackets and the degree of a periodic orbit refers to the Maslov index discussed in chapter \ref{chap:maslov}. In principle, we could consider vector spaces over any field, but in order to avoid issues of orientation and signs, \emph{over the rest of this chapter, it should be assumed that vector fields generated by periodic orbits are taken over $\Z_2$}.

The differential of this chain complex is analogous to the one considered for Morse theory, where the differential of an orbit $x$ is of the form
\begin{equation}
\sum n(x,y) y,
\end{equation}
where the sum is taken over the periodic orbits $y$ satisfying $\mu(y) = \mu(x) - 1$, and $n(x,y)$ is the number of pseudo-holomorphic curves $u(t,s)$ whose limit\footnote{In the $C^\infty$ sense.} as $s \to \infty$ is $y$ and whose limit as $s \to -\infty$ is $x$.

It is a non-trivial fact that this does form a differential, and hence the complex $\CF_*(M,H,J)$ is indeed a chain complex, from which one may construct the homology $\HF_*(M,H,J)$. It is possible to show that this homology does not depend on the chosen Hamiltonian or almost complex structure, and that it coincides with the Morse homology of $M$. For details, see chapter 11 of \cite{audin}.

\section{Filtered Floer Homology}

The notion of Floer homology is, by itself, inadequate for persistence homology, because there is no dependency of the homology on a real variable. The notion of filtered Floer homology is a way to remedy that.

Note that the flow of the `gradient of the action functional' always decreases the value of the action. Therefore, for any cycle $x$, $\partial x$ will always be a combination of orbits $y$ with $\AA(y) \leq \AA(x)$. As a consequence, the differential may be restricted to the (space spanned by the) orbits with action less than some fixed $\lambda \in \R$. The resulting chain complex is denoted $\CF_*^\lambda$, and the resulting homology is the so-called \emph{filtered Floer homology} of $M$, which depends on the parameter $\lambda$. Unlike plain Floer homology, different Hamiltonians yield distinct filtered Floer homologies, as the orbits whose actions are below a certain value of $\lambda$ depend explicitly on the chosen Hamiltonian. Somewhat surprisingly, it only depends on the time-one flow of $H$, $\phi_H$.

\begin{definition}
Let $(M,\omega)$ be a compact aspherical symplectic manifold, $H$ a Hamiltonian on $M$ and $J$ a generic almost complex structure. The \emph{filtered Floer homology} of $H$ and $J$, denoted
\begin{equation}
\HF^\lambda_*(M,H,J), \quad \lambda \in \R
\end{equation}
is the homology of the chain complex $\CF^\lambda_*(M,H,J)$, whose vector spaces are
\begin{equation}
\CF^\lambda_k(M,H) = \{\, x \in \CF_k(M,H) \mid \AA(x) < \lambda\,\}
\end{equation}
and whose differential is the restriction of the differential of the unfiltered Floer homology.
\end{definition}

\begin{prop}\label{prop:ffh}
The filtered Floer homology does not depend on the choice of generic almost complex structure, and depends only on the time-one flow of $H$.
\end{prop}

\begin{proof}
See \cite{polterovich}, just before definition 8.2.3. See also \cite{schwarz}, proposition 3.11 and the remark on the following page.
\end{proof}

\begin{definition}
Let $(M,\omega)$ be a compact aspherical symplectic manifold and $H$ a Hamiltonian on $M$. The \emph{filtered Floer homology} of the Hamiltonian diffeomorphism $\phi_H$, denoted
\begin{equation}
\HF^\lambda_*(M,\phi_H), \quad \lambda \in \R
\end{equation}
is the homology of the chain complex $\CF^\lambda_*(M,H)$.
\end{definition}

This tool is adequate for application of persistence homology, as it gives us a collection of fields indexed on a real variable. To define a persistence module, it remains only to find natural maps $\pi_{ts} \colon \HF^t_* \to \HF^s_*$. Fortunately, there is a natural inclusion of chain complexes $\CF^t_* \hookrightarrow \CF^s_*$ for $t \leq s$, which induces the desired maps in homology.

\begin{remark}
In order to define the barcode of a nondegenerate Hamiltonian diffeomorphism, it is first necessary to check that it is of finite type. This is a consequence of the fact that any nondegenerate Hamiltonian diffeomorphism has a finite number of periodic orbits. The actions of these orbits are the only values of $t$ where the filtered Floer homology could change, which proves part \ref{pm1} of the definition of finite type. Semicontinuity (part \ref{pm2}) is also a consequence of this fact. Finally, \ref{pm3} can be shown by considering values of $t$ less than minimal action of the orbits.
\end{remark}

\begin{remark}
While proposition \ref{prop:ffh} shows that the spaces of the filtered Floer homology do not depend on the Hamiltonian $H$ chosen to represent a given Hamiltonian diffeomorphism, it does not prove that the maps $\pi_{st}$ do not. We have been unable to find a proof of this fact, but \cite{polterovich} tacitly assumes it, so we take as given that it is true. Regardless, the conclusions of this paper do not rely on this fact.
\end{remark}

\begin{definition}
Let $(M,\omega)$ be a compact aspherical symplectic manifold and $\phi_H$ a Hamiltonian diffeomorphism on $M$. The \emph{barcode of $\phi_H$ in degree $*$}, denoted $\BB_*(\phi_H)$, is defined as the barcode associated to the persistence module given by the filtered Floer homology of $\phi_H$ in degree $*$.

The \emph{total barcode of $\phi_H$}, denoted $\BB(\phi_H)$, is the union $\bigcup_{* \in \Z} \BB_*(\phi_H)$.
\end{definition}

The study of what properties of a Hamiltonian diffeomorphism can be deduced from its barcode is an active area of research: see for example \cite{polterovich}, \cite{kislev2022bounds}, \cite{roux2018barcodes}, and \cite{polterovich2016autonomous}. In this thesis, we will instead compute the barcodes of a few such diffeomorphisms, namely of non-autonomous Hamiltonians, as the barcodes computed in the literature are mostly those of small autonomous Hamiltonians, in which case they can be computed instead via filtered Morse homology (see chapter \ref{subsecautonomous}).